---
title: "Practice Exercises: DataFrames and Data Processing in R"
author: "Einat Borohovich"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Base Datasets

These datasets are provided for the exercises. Run this chunk first!

```{r base-data}
# Sample patient data with lab values (for examples)
set.seed(123)
patient_data <- data.frame(
  patient_id = 101:150,
  age = sample(45:85, 50, replace = TRUE),
  creatinine = round(rnorm(50, mean = 1.2, sd = 0.4), 2),
  egfr = round(rnorm(50, mean = 65, sd = 20), 1),
  gender = sample(c("M", "F"), 50, replace = TRUE)
)

# Sample diabetes patient data (for examples)
diabetes_patients <- data.frame(
  patient_id = 201:250,
  diabetes_type = sample(c("Type 1", "Type 2", "Prediabetes"), 50, 
                         replace = TRUE, prob = c(0.15, 0.65, 0.20)),
  hba1c = round(rnorm(50, mean = 7.5, sd = 1.5), 1),
  glucose = round(rnorm(50, mean = 140, sd = 30), 0),
  age = sample(35:75, 50, replace = TRUE)
)
```

---

# Instructions

**These are practice exercises to help you:**
- Build confidence working with clinical data in R
- Practice vectorization techniques that make analysis faster
- Learn essential data manipulation packages

💡 **Tips:**
- Try to solve each exercise on your own first
- If you get stuck, check the solution file
- Run each code chunk to see your results (Ctrl+Enter or Cmd+Enter)
- Feel free to experiment and modify the code!

---

# Exercise 1: Creating and Exploring Patient DataFrames

You are analyzing patient data from a diabetes clinic study.

## 1.1 Create a Patient DataFrame

Create a dataframe called `patients` with the following information for 6 patients:

- `patient_id`: 101, 102, 103, 104, 105, 106
- `age`: 45, 32, 67, 54, 28, 71
- `gender`: "M", "F", "M", "F", "M", "F"
- `systolic_bp`: 120, 118, 145, 135, 110, 150
- `has_diabetes`: TRUE, FALSE, TRUE, FALSE, FALSE, TRUE

```{r create-dataframe}
# Your code here


```

## 1.2 Explore Your Data

Write code to:

a) Display the structure of the dataframe (use `str()`)
b) Show the first 3 rows (use `head()`)
c) Calculate the mean age of all patients
d) Count how many patients have diabetes

```{r explore-dataframe}
# a) Structure


# b) First 3 rows


# c) Mean age


# d) Count diabetes patients


```

**Clinical insight:** These basic exploration commands are essential when you first receive a dataset. Always check structure and summary statistics before analysis!

---

# Exercise 2: Why Vectorization Matters for Clinical Data

## 2.1 Speed Comparison: Temperature Conversion

**Your task:** Complete the temperature conversion code to compare loop vs vectorized approaches.

Formula: °F = (°C × 9/5) + 32

```{r temperature-conversion}
# Set seed for reproducibility
set.seed(123)

# Create vector of 100 million temperatures to show clear timing difference
n <- 100000000
celsius <- runif(n, min = 35, max = 40)

# METHOD 1: Using a loop (slow - DON'T do this!)
fahrenheit_loop <- numeric(n)
time_loop <- system.time({
  for(i in 1:n) {
    fahrenheit_loop[i] <- # Your code: convert celsius[i] to fahrenheit
  }
})

# METHOD 2: Vectorized operation (fast - ALWAYS do this!)
time_vector <- system.time({
  fahrenheit_vector <- # Your code: convert entire celsius vector at once
})

# Results with high precision
cat("Loop time:", sprintf("%.6f", time_loop[3]), "seconds\n")
cat("Vectorized time:", sprintf("%.6f", time_vector[3]), "seconds\n")
if(time_vector[3] > 0) {
  cat("Speedup:", sprintf("%.1fx", time_loop[3] / time_vector[3]), "faster\n")
} else {
  cat("Vectorized operation is significantly faster!\n")
}

# Verify both give same results (check first 1000 to save time)
all.equal(fahrenheit_loop[1:1000], fahrenheit_vector[1:1000])
```

**What you should learn:** The vectorized version is much faster! This difference matters with real clinical datasets.

💡 **Key takeaway:** ALWAYS vectorize. Never write loops for simple operations.

## 2.2 BMI Classification - Practical Vectorization

You have patient weights and need to classify them. Do this using vectorized operations (NO LOOPS!).

```{r weight-analysis}
# Patient weights in kg
weights_kg <- c(55, 72, 68, 95, 58, 81, 104, 63, 77, 89)

# Task 1: Convert to pounds (1 kg = 2.20462 lbs)
weights_lbs <- # Your code here

# Task 2: Categorize as:
# "Underweight" (<60 kg), "Normal" (60-90 kg), "Overweight" (>90 kg)
weight_category <- # Your code here (hint: use ifelse)

# Display results in a nice table
results <- data.frame(
  kg = weights_kg,
  lbs = round(weights_lbs, 1),
  category = weight_category
)
print(results)
```

---

# Exercise 3: Using dplyr for Clinical Data Analysis

`dplyr` has 5 main "verbs" for data manipulation: filter, select, mutate, arrange, summarize. Let's practice them!

## 3.1 Filter and Select

We'll use the built-in `mtcars` dataset. In clinical work, you'd use the same techniques on patient data.

**Your task:** Find efficient cars (mpg > 20) with good power (hp > 100), show only relevant columns.

```{r dplyr-filter}
# Load dplyr
library(dplyr)

# Use the pipe operator %>% to chain operations:
result <- mtcars %>%
  filter(___________) %>%      # Filter: mpg > 20 AND hp > 100
  select(___________) %>%      # Select: mpg, hp, wt
  arrange(_________)           # Arrange: by mpg, descending

# Display result
print(result)
```

💡 **Clinical example:** Filter patients by `age > 65 & creatinine > 1.5`, select lab values, arrange by risk.

## 3.2 Create New Columns with mutate()

**Task:** Add a BMI column to the patient data.

```{r dplyr-mutate}
# First add weight and height columns to the patients dataframe
patients_with_data <- patients %>%
  mutate(
    weight = c(___________________),      # Add: 70, 65, 80, 72, 58, 85
    height_m = c(_________________)       # Add: 1.75, 1.68, 1.82, 1.70, 1.65, 1.78
  )

# Now calculate BMI
patients_with_bmi <- patients_with_data %>%
  mutate(
    bmi = _________  # Calculate: weight / (height_m^2)
  )

print(patients_with_bmi)
```

## 3.3 Group and Summarize

**Clinical scenario:** Calculate average values by patient groups.

```{r dplyr-summarize}
# Calculate mean mpg and hp for each cylinder group
summary_stats <- mtcars %>%
  group_by(_______) %>%           # Group by: cyl
  summarize(
    avg_mpg = _________,          # mean of mpg, rounded to 2 decimals
    avg_hp = _________            # mean of hp, rounded to 2 decimals
  )

print(summary_stats)
```

💡 **Clinical translation:** Same technique works for "calculate mean HbA1c by diabetes type" or "average BP by age group"

---

# Exercise 4: Reshaping Data with tidyr

**Clinical scenario:** Blood pressure data often comes in "wide" format (one row per patient, columns for each visit). For analysis, we need "long" format (one row per measurement).

```{r tidyr-exercise}
library(tidyr)

# Blood pressure data - wide format (typical EMR export)
bp_wide <- data.frame(
  patient_id = c(1, 2, 3),
  visit_1 = c(120, 135, 118),
  visit_2 = c(122, 138, 116),
  visit_3 = c(119, 140, 115)
)

print("Wide format:")
print(bp_wide)

# Convert to long format
# Final columns should be: patient_id, visit, systolic_bp
bp_long <- bp_wide %>%
  pivot_longer(
    cols = ___________,           # Which columns to pivot? (hint: starts_with("visit_"))
    names_to = "visit",            # Name for the new "visit" column
    values_to = "systolic_bp"      # Name for the values column
  )

print("Long format:")
print(bp_long)
```

💡 **Why this matters:** Long format is essential for:
- Plotting trends over time
- Statistical models (mixed models, GEE)
- Calculating changes between visits

---

# Exercise 5: Visualizing Data with ggplot2

Create a visualization of the patient data using ggplot2.

```{r visualization}
library(ggplot2)

# First, add weight, height, and BMI to the patients dataframe
# Use mutate() to add these columns:
patients_complete <- patients %>%
  mutate(
    weight = c(___________________),      # Add: 70, 65, 80, 72, 58, 85
    height_m = c(_________________),      # Add: 1.75, 1.68, 1.82, 1.70, 1.65, 1.78
    bmi = _________                       # Calculate: weight / (height_m^2)
  )

# Create a scatter plot: Age vs BMI, colored by diabetes status
ggplot(patients_complete, aes(x = _______, y = _______, color = _______)) +
  geom_point(size = 3) +
  labs(
    title = "BMI vs Age by Diabetes Status",
    x = "Age (years)",
    y = "BMI (kg/m²)",
    color = "Diabetes"
  ) +
  theme_minimal()
```

💡 **ggplot2 syntax:** `ggplot(data, aes(x, y, color)) + geom_point()`

---

# Exercise 6: Complete Analysis Pipeline

Combine everything you've learned! Using the `patients` dataframe from Exercise 1:

**Task 1:** Add a blood pressure category column
**Task 2:** Calculate statistics by category

```{r combined-analysis}
# Task 1: Categorize blood pressure
# "Normal" (<120), "Elevated" (120-139), "Hypertension" (≥140)
patients$bp_category <- ifelse(
  ________,      # Condition for "Normal"
  "Normal",
  ifelse(
    ________,    # Condition for "Elevated"  
    "Elevated",
    "Hypertension"
  )
)

print(patients)

# Task 2: Calculate average BP by category
patients %>%
  group_by(________) %>%
  summarize(
    n_patients = ________,              # Count patients (use n())
    avg_bp = ________                   # Average systolic_bp
  )
```

---

# Ready to Check Your Work?

✅ Run each code chunk to test your solutions
✅ Compare with the solution file when you're done
✅ Try modifying the code to experiment with different values

**Remember:** The goal is learning, not perfection. If something doesn't work, that's a learning opportunity!
