---
title: "Practice Exercises - SOLUTIONS"
author: "Einat Borohovich"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: readable
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Base Datasets

These datasets are used throughout the examples and solutions.

```{r base-data}
# Sample patient data with lab values
set.seed(123)
patient_data <- data.frame(
  patient_id = 101:150,
  age = sample(45:85, 50, replace = TRUE),
  creatinine = round(rnorm(50, mean = 1.2, sd = 0.4), 2),
  egfr = round(rnorm(50, mean = 65, sd = 20), 1),
  gender = sample(c("M", "F"), 50, replace = TRUE)
)

# Sample diabetes patient data
diabetes_patients <- data.frame(
  patient_id = 201:250,
  diabetes_type = sample(c("Type 1", "Type 2", "Prediabetes"), 50, 
                         replace = TRUE, prob = c(0.15, 0.65, 0.20)),
  hba1c = round(rnorm(50, mean = 7.5, sd = 1.5), 1),
  glucose = round(rnorm(50, mean = 140, sd = 30), 0),
  age = sample(35:75, 50, replace = TRUE)
)
```

---

# Teaching Notes

This solution guide provides:
- Complete working code for all exercises
- Clinical context and explanations
- Alternative approaches where applicable

---

# Exercise 1: Creating and Exploring Patient DataFrames

## 1.1 Create a Patient DataFrame

**Solution:**
```{r create-dataframe-solution}
# Create the patients dataframe
patients <- data.frame(
  patient_id = c(101, 102, 103, 104, 105, 106),
  age = c(45, 32, 67, 54, 28, 71),
  gender = c("M", "F", "M", "F", "M", "F"),
  systolic_bp = c(120, 118, 145, 135, 110, 150),
  has_diabetes = c(TRUE, FALSE, TRUE, FALSE, FALSE, TRUE)
)

print(patients)
```

**Teaching points:**
- Column names should be informative (`systolic_bp` not just `bp`)
- Logical vectors (TRUE/FALSE) are perfect for binary clinical variables
- Using `c()` to concatenate values is fundamental

## 1.2 Explore Your Data

**Solution (matching homework):**
```{r explore-dataframe-solution}
# a) Structure
str(patients)

# b) First 3 rows
head(patients, 3)

# c) Mean age
mean(patients$age)

# d) Count diabetes patients
sum(patients$has_diabetes)
```

**Enhanced output (with better formatting):**
```{r explore-enhanced}
# Same commands but with nice output
cat("Mean age:", mean(patients$age), "years\n")
cat("Patients with diabetes:", sum(patients$has_diabetes), "out of", nrow(patients), "\n")
```

**Alternative solutions students might use:**
```{r alternatives-1}
# For counting diabetes patients:
table(patients$has_diabetes)  # Shows both TRUE and FALSE counts
nrow(patients[patients$has_diabetes == TRUE, ])  # Works but verbose
length(which(patients$has_diabetes))  # Also works
```

**Common mistakes:**
- Forgetting the `$` when accessing columns
- Using `length()` instead of `nrow()` for counting rows
- Not understanding that `sum()` on logical vectors counts TRUE values

---

# Exercise 2: Why Vectorization Matters for Clinical Data

## 2.1 Speed Comparison: Temperature Conversion

**Solution:**
```{r temperature-conversion-solution}
set.seed(123)

# Use 100 million values to show clear timing difference
n <- 100000000
celsius <- runif(n, min = 35, max = 40)

# METHOD 1: Loop (slow)
fahrenheit_loop <- numeric(n)
time_loop <- system.time({
  for(i in 1:n) {
    fahrenheit_loop[i] <- (celsius[i] * 9/5) + 32
  }
})

# METHOD 2: Vectorized (fast)
time_vector <- system.time({
  fahrenheit_vector <- (celsius * 9/5) + 32
})

# Results with high precision
cat("Loop time:", sprintf("%.6f", time_loop[3]), "seconds\n")
cat("Vectorized time:", sprintf("%.6f", time_vector[3]), "seconds\n")
if(time_vector[3] > 0) {
  cat("Speedup:", sprintf("%.1fx", time_loop[3] / time_vector[3]), "faster\n")
} else {
  cat("Vectorized operation is significantly faster!\n")
}

# Verify identical results (check first 1000 to save time)
all.equal(fahrenheit_loop[1:1000], fahrenheit_vector[1:1000])
```

**Teaching points:**
- Formula: `(celsius * 9/5) + 32`
- Loop version: `fahrenheit_loop[i] <- (celsius[i] * 9/5) + 32`
- Vectorized version: `fahrenheit_vector <- (celsius * 9/5) + 32`
- The vectorized version is typically 50-100x faster
- With 100 million values, loop takes many seconds, vectorized takes < 1 second
- This speed difference is critical when working with large clinical datasets

## 2.2 BMI Classification - Practical Vectorization

**Solution:**
```{r weight-analysis-solution}
weights_kg <- c(55, 72, 68, 95, 58, 81, 104, 63, 77, 89)

# Convert to pounds
weights_lbs <- weights_kg * 2.20462

# Categorize using nested ifelse (most common approach)
weight_category <- ifelse(weights_kg < 60, "Underweight",
                         ifelse(weights_kg <= 90, "Normal", "Overweight"))

# Display results
results <- data.frame(
  kg = weights_kg,
  lbs = round(weights_lbs, 1),
  category = weight_category
)
print(results)
```

**Teaching points:**
- `ifelse()` is vectorized - it processes all values at once
- Nested `ifelse()` for multiple categories is the standard approach you learned in class

---

# Exercise 3: Using dplyr for Clinical Data Analysis

## 3.1 Filter, Select, and Sort

**Solution:**
```{r dplyr-filter-solution, message=FALSE}
library(dplyr)

result <- mtcars %>%
  filter(mpg > 20 & hp > 100) %>%
  select(mpg, hp, wt) %>%
  arrange(desc(mpg))

print(result)
```

**Alternative without pipe:**
```{r no-pipe}
# Same result without %>%
result_no_pipe <- arrange(
  select(
    filter(mtcars, mpg > 20 & hp > 100),
    mpg, hp, wt
  ),
  desc(mpg)
)
```

**Clinical translation example:**
```{r clinical-example}
# Using the patient_data we created earlier
patient_results <- patient_data %>%
  filter(age > 65 & creatinine > 1.5) %>%  # High-risk elderly
  select(patient_id, age, creatinine, egfr) %>%
  arrange(desc(creatinine))  # Highest risk first

print(patient_results)
```



## 3.2 Create New Columns with mutate()

**Solution:**
```{r dplyr-mutate-solution}
# First add weight and height columns
patients_with_data <- patients %>%
  mutate(
    weight = c(70, 65, 80, 72, 58, 85),
    height_m = c(1.75, 1.68, 1.82, 1.70, 1.65, 1.78)
  )

# Now calculate BMI
patients_with_bmi <- patients_with_data %>%
  mutate(
    bmi = weight / (height_m^2)
  )

print(patients_with_bmi)
```

**Alternative - combine in one step:**
```{r mutate-alternative}
# Can also do it all at once
patients_with_bmi_v2 <- patients %>%
  mutate(
    weight = c(70, 65, 80, 72, 58, 85),
    height_m = c(1.75, 1.68, 1.82, 1.70, 1.65, 1.78),
    bmi = weight / (height_m^2)  # Can use columns created in same mutate()
  )

print(patients_with_bmi_v2)
```

**Teaching points:**
- `mutate()` creates new columns or modifies existing ones
- Can create multiple columns in one `mutate()` call
- New columns can use other columns created in the same `mutate()`
- BMI formula: weight (kg) / height (m)²

## 3.3 Group and Summarize Lab Results

**Solution:**
```{r dplyr-summarize-solution}
summary_stats <- mtcars %>%
  group_by(cyl) %>%
  summarize(
    avg_mpg = round(mean(mpg), 2),
    avg_hp = round(mean(hp), 2)
  )

print(summary_stats)
```

**Alternative rounding approaches:**
```{r rounding-alternatives}
# Method 1: Round after calculation (shown above)

# Method 2: Round all at once
mtcars %>%
  group_by(cyl) %>%
  summarize(avg_mpg = mean(mpg), avg_hp = mean(hp)) %>%
  mutate(across(where(is.numeric), ~round(., 2)))

# Method 3: Using format for display only
mtcars %>%
  group_by(cyl) %>%
  summarize(avg_mpg = mean(mpg), avg_hp = mean(hp)) %>%
  mutate(across(where(is.numeric), ~sprintf("%.2f", .)))
```

**Clinical example:**
```{r clinical-group}
# Calculate mean HbA1c by diabetes type using the diabetes_patients data
diabetes_summary <- diabetes_patients %>%
  group_by(diabetes_type) %>%
  summarize(
    n_patients = n(),
    mean_hba1c = round(mean(hba1c, na.rm = TRUE), 1),
    mean_glucose = round(mean(glucose, na.rm = TRUE), 1)
  )

print(diabetes_summary)
```

---

# Exercise 4: Reshaping Data with tidyr

**Solution:**
```{r tidyr-solution}
library(tidyr)

bp_wide <- data.frame(
  patient_id = c(1, 2, 3),
  visit_1 = c(120, 135, 118),
  visit_2 = c(122, 138, 116),
  visit_3 = c(119, 140, 115)
)

# Convert to long format (matching class example)
bp_long <- bp_wide %>%
  pivot_longer(
    cols = starts_with("visit_"),
    names_to = "visit",
    values_to = "systolic_bp"
  )

print(bp_long)
```

**Note:** The visit column will contain "visit_1", "visit_2", "visit_3". This matches what was taught in class.

**Why long format matters - demonstration:**
```{r long-format-benefits}
# Now we can easily plot trends
library(ggplot2)

ggplot(bp_long, aes(x = visit, y = systolic_bp, group = patient_id, color = factor(patient_id))) +
  geom_line() +
  geom_point() +
  labs(title = "Blood Pressure Trends by Patient",
       x = "Visit", y = "Systolic BP (mmHg)",
       color = "Patient ID") +
  theme_minimal()

# Or calculate change from baseline
bp_long %>%
  group_by(patient_id) %>%
  mutate(change_from_baseline = systolic_bp - first(systolic_bp))
```

**Teaching points:**
- Long format is essential for: time series plots, mixed models, calculating changes
- Wide format is better for: viewing, data entry, simple comparisons
- Most statistical functions expect long format

---

# Exercise 5: Visualizing Data with ggplot2

**Solution:**
```{r visualization-solution}
library(ggplot2)

# Add BMI to patients
patients_complete <- patients %>%
  mutate(
    weight = c(70, 65, 80, 72, 58, 85),
    height_m = c(1.75, 1.68, 1.82, 1.70, 1.65, 1.78),
    bmi = weight / (height_m^2)
  )

# Create scatter plot
ggplot(patients_complete, aes(x = age, y = bmi, color = has_diabetes)) +
  geom_point(size = 3) +
  labs(
    title = "BMI vs Age by Diabetes Status",
    x = "Age (years)",
    y = "BMI (kg/m²)",
    color = "Diabetes"
  ) +
  theme_minimal()
```

**Alternative with better colors:**
```{r viz-enhanced}
ggplot(patients_complete, aes(x = age, y = bmi, color = has_diabetes)) +
  geom_point(size = 4, alpha = 0.7) +
  scale_color_manual(values = c("FALSE" = "#2ecc71", "TRUE" = "#e74c3c"),
                     labels = c("No", "Yes")) +
  labs(
    title = "BMI vs Age by Diabetes Status",
    x = "Age (years)",
    y = "BMI (kg/m²)",
    color = "Has Diabetes"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))
```

**Teaching points:**
- Basic ggplot structure: `ggplot(data, aes()) + geom_*()`
- `aes()` maps data to visual properties (x, y, color, size)
- Add layers with `+`
- `theme_minimal()` gives clean appearance

---

# Exercise 6: Complete Analysis Pipeline

**Solution:**
```{r combined-solution}
# Task 1: Categorize blood pressure
patients$bp_category <- ifelse(patients$systolic_bp < 120, "Normal",
                              ifelse(patients$systolic_bp < 140, "Elevated", "Hypertension"))

print(patients)

# Task 2: Calculate average BP by category (matching homework)
patients %>%
  group_by(bp_category) %>%
  summarize(
    n_patients = n(),
    avg_bp = round(mean(systolic_bp), 1)
  )
```

**Enhanced version with additional statistics:**
```{r combined-enhanced}
# You can add more summary statistics for deeper analysis
bp_summary <- patients %>%
  group_by(bp_category) %>%
  summarize(
    n_patients = n(),
    avg_bp = round(mean(systolic_bp), 1),
    min_bp = min(systolic_bp),
    max_bp = max(systolic_bp)
  ) %>%
  arrange(desc(avg_bp))  # Order by severity

print(bp_summary)
```

**Enhanced visualization:**
```{r combined-viz}
# Visualize the distribution
library(ggplot2)

ggplot(patients, aes(x = bp_category, y = systolic_bp, fill = bp_category)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, alpha = 0.5) +
  scale_fill_manual(values = c("Normal" = "#2ecc71", 
                                "Elevated" = "#f39c12", 
                                "Hypertension" = "#e74c3c")) +
  labs(title = "Blood Pressure Distribution by Category",
       x = "BP Category", y = "Systolic BP (mmHg)") +
  theme_minimal() +
  theme(legend.position = "none")
```

**Alternative categorization using case_when:**
```{r case-when-alternative}
# Using case_when (cleaner for multiple conditions)
patients <- patients %>%
  mutate(bp_category_v2 = case_when(
    systolic_bp < 120 ~ "Normal",
    systolic_bp < 140 ~ "Elevated",
    TRUE ~ "Hypertension"
  ))
```

---

# Additional Teaching Examples

## Example 1: Complete Clinical Workflow

```{r complete-workflow}
# Simulate patient data
set.seed(456)
n_patients <- 100

clinical_data <- data.frame(
  patient_id = 1:n_patients,
  age = sample(25:85, n_patients, replace = TRUE),
  gender = sample(c("M", "F"), n_patients, replace = TRUE),
  systolic_bp = round(rnorm(n_patients, mean = 130, sd = 15)),
  diastolic_bp = round(rnorm(n_patients, mean = 80, sd = 10)),
  glucose = round(rnorm(n_patients, mean = 100, sd = 20)),
  has_diabetes = sample(c(TRUE, FALSE), n_patients, replace = TRUE, prob = c(0.15, 0.85))
)

# Complete analysis pipeline
analysis_results <- clinical_data %>%
  # Add computed variables
  mutate(
    bp_category = case_when(
      systolic_bp < 120 ~ "Normal",
      systolic_bp < 140 ~ "Elevated",
      TRUE ~ "Hypertension"
    ),
    glucose_category = ifelse(glucose < 100, "Normal", "Impaired")
  ) %>%
  # Filter to high-risk patients
  filter(systolic_bp >= 140 | glucose >= 126) %>%
  # Group analysis
  group_by(bp_category, has_diabetes) %>%
  summarize(
    n = n(),
    mean_age = round(mean(age), 1),
    mean_glucose = round(mean(glucose), 1),
    .groups = "drop"
  ) %>%
  arrange(desc(mean_glucose))

print(analysis_results)
```

## Example 2: Time Series Data

```{r time-series}
# Patient lab values over time
lab_data <- data.frame(
  patient_id = rep(1:3, each = 4),
  date = rep(seq(as.Date("2025-01-01"), by = "1 week", length.out = 4), 3),
  creatinine = c(1.1, 1.3, 1.5, 1.4,  # Patient 1
                 0.9, 0.9, 1.0, 0.9,  # Patient 2
                 1.8, 2.1, 2.3, 2.5)  # Patient 3
)

# Calculate changes from baseline
lab_trends <- lab_data %>%
  group_by(patient_id) %>%
  arrange(date) %>%
  mutate(
    baseline = first(creatinine),
    change = creatinine - baseline,
    pct_change = round((change / baseline) * 100, 1)
  ) %>%
  ungroup()

print(lab_trends)

# Identify worsening cases
worsening <- lab_trends %>%
  filter(pct_change > 20) %>%
  select(patient_id, date, creatinine, pct_change)

print("Patients with >20% increase in creatinine:")
print(worsening)
```

---

# Common Student Questions & Answers

**Q: When should I use %>% vs regular function calls?**
A: Use %>% when you have multiple steps. If you're only doing one operation, either is fine. The pipe makes code more readable when you have 3+ operations.

**Q: What's the difference between filter() and subset()?**
A: `filter()` is from dplyr and works better with pipes. `subset()` is base R. Use `filter()` for consistency with dplyr.

**Q: Why does vectorization matter if my dataset is small?**
A: Two reasons: (1) It's good practice for when you work with larger data, (2) Even with small data, it makes code cleaner and more readable.

**Q: Should I always use long format?**
A: No - use long for: plotting, modeling, calculating changes. Use wide for: viewing data, entering data, simple reports.

**Q: What if I have missing values (NA)?**
A: Most functions have `na.rm = TRUE` parameter:
```{r na-example}
values_with_na <- c(120, 130, NA, 140, 125)
mean(values_with_na)  # Returns NA
mean(values_with_na, na.rm = TRUE)  # Returns 128.75
```

---

# Summary of Key Concepts

1. **Vectorization > Loops**: Always prefer vectorized operations
2. **dplyr verbs**: filter, select, mutate, group_by, summarize
3. **Pipe operator %>%**: Chains operations left to right
4. **Data shapes**: Wide for viewing, long for analysis
5. **Functions**: Reusable code for repeated tasks

