---
title: "DataFrames & Packages in R"
subtitle: "Mastering Vectorization and Modern R Tools"
author: "R Workshop"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: cosmo
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 7,
  fig.height = 4
)
```

# Part 1: DataFrames Basics

## What is a DataFrame?

A DataFrame is a table with rows and columns. Each column can be a different type, but all must have the same length.

```{r dataframe-creation}
# Create a DataFrame
df <- data.frame(
  name = c("Alice", "Bob", "Carol"),
  age = c(25, 30, 35),
  score = c(85.5, 92.3, 78.9)
)

print(df)
```

## Exploring DataFrames

```{r dataframe-explore}
# View structure
str(df)

# Access columns
df$name
df[["age"]]

# Access rows
df[1, ]  # First row
```

---

# Part 2: Vectorization vs Loops

## Why Vectorize?

Vectorization applies operations to entire vectors at once - it's **10-100x faster** than loops.

```{r vectorization-demo}
n <- 100000000
x <- 1:n
# Loop approach (slow)
print(system.time({
  result_loop <- numeric(n)
  for (i in 1:n) {
    result_loop[i] <- sqrt(x[i])
  }
}), digits = 7)
# Vectorized approach (fast)
print(system.time({
  result_vec <- sqrt(x)
}), digits = 7)
```

## Common Vectorized Functions

```{r vectorized-functions}
x <- c(1, 2, 3, 4, 5)

# Numeric
x + 10
x^2
sqrt(x)

# Logical
x > 3
x == 4

# Conditional
ifelse(x > 3, "High", "Low")
```

---

# Part 3: Essential Packages

## Setup

```{r load-packages}
library(dplyr)
library(tidyr)
library(ggplot2)
```

## Sample Data

```{r sample-data}
patients <- data.frame(
  name = c("Alice", "Bob", "Carol", "David", "Eve"),
  age = c(25, 30, 35, 40, 28),
  weight = c(65, 80, 70, 85, 62),
  height = c(165, 180, 170, 175, 160),
  treatment = c("A", "B", "A", "B", "A")
)

patients
```

## dplyr: The 5 Verbs

### filter() - Subset rows

```{r filter}

patients %>%
  filter(age > 30)
```

### select() - Choose columns

```{r select}
patients %>%
  select(name, age, treatment)
```

### mutate() - Create columns

```{r mutate}
patients %>%
  mutate(bmi = weight / (height/100)^2)
```

### arrange() - Sort rows

```{r arrange}
patients %>%
  arrange(desc(age))
```

### summarize() - Aggregate data

```{r summarize}
patients %>%
  group_by(treatment) %>%
  summarize(
    mean_age = mean(age),
    count = n()
  )
```

## Complete Pipeline with Pipes

```{r pipeline}
# Clean workflow using %>%
results <- patients %>%
  filter(age > 25) %>%
  mutate(bmi = weight / (height/100)^2) %>%
  group_by(treatment) %>%
  summarize(avg_bmi = mean(bmi))

print(results)
```

## tidyr: Reshape Data

```{r tidyr}
# Wide to Long
wide_data <- data.frame(
  patient = c("Alice", "Bob"),
  week1 = c(120, 130),
  week2 = c(118, 128)
)

wide_data

long_data <- wide_data %>%
  pivot_longer(
    cols = starts_with("week"),
    names_to = "timepoint",
    values_to = "bp"
  )

print(long_data)
```

**Pro Tip:** Long format works best with ggplot2 and group_by().

## data.table: For Big Data

```{r datatable}
library(data.table)

dt <- as.data.table(patients)

# Syntax: DT[i, j, by]
# i = rows, j = columns, by = grouping

dt[age > 30, .(name, age)]
dt[, mean(age), by = treatment]
```

**Use data.table when:** datasets > 1GB or speed is critical.

## Other Essential Packages

```{r other-packages, eval=FALSE}


# lubridate - Dates
library(lubridate)
ymd("2024-01-15")

# stringr - Strings
library(stringr)
str_to_upper("hello")


df <- your_data_frame
ggplot(df, aes(x, y))

# Or use a different name like:
# This causes the error:
ggplot(data, aes(x = bmi, y = age))

# Should be:
patients_bmi <- patients %>%
  mutate(bmi = weight / (height/100)^2)

ggplot(patients, aes(x = age, y = weight)) +
  geom_point()


# ggplot2 - Visualization
patients_bmi <- patients %>%
  mutate(bmi = weight / (height/100)^2)

ggplot(patients_bmi, aes(x = age, y = bmi, color = treatment, size = weight)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "BMI vs Age by Treatment Group",
    x = "Age (years)",
    y = "BMI (kg/m²)",
    color = "Treatment"
  ) +
  theme_minimal()
```

---

# Part 4: Practical Example

## Patient Data Analysis

```{r practical-example}
# Step 1: Create sample data
set.seed(123)
patient_data <- data.frame(
  id = 1:50,
  age = sample(25:65, 50, replace = TRUE),
  weight = rnorm(50, mean = 75, sd = 10),
  height = rnorm(50, mean = 170, sd = 8),
  treatment = sample(c("A", "B"), 50, replace = TRUE),
  outcome = sample(c("Improved", "No Change"), 50, 
                   replace = TRUE, prob = c(0.7, 0.3))
)

# Step 2: Clean and transform
patient_clean <- patient_data %>%
  mutate(
    bmi = weight / (height/100)^2,
    bmi_cat = ifelse(bmi < 25, "Normal", "Overweight")
  )

# Step 3: Analyze
summary_stats <- patient_clean %>%
  group_by(treatment) %>%
  summarize(
    n = n(),
    mean_bmi = mean(bmi),
    pct_improved = mean(outcome == "Improved") * 100
  )

print(summary_stats)

# Step 4: Visualize
ggplot(patient_clean, aes(x = treatment, y = bmi, fill = treatment)) +
  geom_boxplot() +
  labs(title = "BMI by Treatment Group") +
  theme_minimal()
```

---

# Best Practices

## DO ✓

```{r do-practices, eval=FALSE}
# Always vectorize
x^2 + sqrt(x)  # Good

# Pre-allocate if loops needed
# Calculate running total where each step depends on previous result
n <- 5
x <- c(10, 5, 3, 7, 2)

# Pre-allocate result vector
result <- numeric(n)
result[1] <- x[1]  # First value

# Each value = previous result + current x
for (i in 2:n) {
  result[i] <- result[i-1] + x[i]
}

print(result)  # 10, 15, 18, 25, 27

# Use data.table for big data
library(data.table)
```

## DON'T ✗

```{r dont-practices, eval=FALSE}
# Don't loop when you can vectorize
for (i in 1:n) result[i] <- x[i] * 2  # Bad
result <- x * 2  # Good

# Don't grow vectors
result <- c()
for (i in 1:n) result <- c(result, x[i])  # Very slow!
```

---

# Key Takeaways

- **Vectorization is 10-100x faster** than loops
- **dplyr** for readable code, **data.table** for big data
- **Pipes (`%>%`)** create clear workflows
- **tidyverse** handles 90% of data tasks

---

# Resources

**Book:** R for Data Science - https://r4ds.had.co.nz/

**Documentation:**
- dplyr: https://dplyr.tidyverse.org/
- data.table: https://rdatatable.gitlab.io/data.table/

**Cheatsheets:** https://www.rstudio.com/resources/cheatsheets/

**Community:**
- RStudio Community: https://community.rstudio.com
- Stack Overflow: [r] tag

---

**Happy Coding with R!**
