---
title: "Data Visualization — Practice Solutions"
subtitle: "R for Clinicians · Lecture 07"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)
```

# About this file

One worked solution per problem. Yours may differ — there's no single
correct chart. What matters is that the chart applies the principles from
the lecture deliberately.

## Setup

```{r setup-data}
library(tidyverse)
library(mlbench)
library(scales)

data("PimaIndiansDiabetes")

pima <- PimaIndiansDiabetes |>
  as_tibble() |>
  filter(glucose > 0, mass > 0, pressure > 0)
```

---

# Problem 1 — BMI categories

The point of this problem: `fct_reorder` is a tool, not a reflex. BMI
categories have an inherent order (Underweight → Obese III). Sorting by
count would scramble that order and force the reader to mentally
reconstruct the natural sequence. The order of the x-axis is itself
information.

```{r p1-solution}
pima_bmi <- pima |>
  mutate(
    bmi_class = case_when(
      mass < 18.5  ~ "Underweight",
      mass < 25.0  ~ "Normal",
      mass < 30.0  ~ "Overweight",
      mass < 35.0  ~ "Obese I",
      mass < 40.0  ~ "Obese II",
      TRUE         ~ "Obese III"
    ),
    bmi_class = factor(bmi_class,
                       levels = c("Underweight", "Normal", "Overweight",
                                  "Obese I", "Obese II", "Obese III"))
  )

ggplot(pima_bmi, aes(x = bmi_class)) +
  geom_bar(fill = "#d97757") +
  scale_y_continuous(expand = expansion(mult = c(0, .05))) +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor   = element_blank()) +
  labs(
    title    = "Most patients in the cohort are overweight or obese",
    subtitle = "WHO BMI classification, n = 752",
    x        = NULL,
    y        = "Number of patients"
  )
```

The `factor()` call with explicit `levels` is what locks the natural
order. Without it, the categories appear alphabetically (Normal, Obese I,
Obese II, Obese III, Overweight, Underweight) — which is nonsense.

---

# Problem 2 — Bin width

Same data, three completely different stories. `pedigree` ranges roughly
0.08 to 2.4 with a strong right skew. `binwidth = 0.3` collapses the
skew, `binwidth = 0.05` fragments the shape into noise, `binwidth = 0.1`
is the sweet spot.

```{r p2-too-fine, fig.height=3.5}
ggplot(pima, aes(x = pedigree)) +
  geom_histogram(binwidth = 0.05, fill = "#c45e5e", color = "white") +
  theme_minimal() +
  labs(title = "binwidth = 0.05 — too fine",
       x = "Pedigree score", y = "Count")
```

Bars start oscillating because of small-sample variation within each
narrow bin. Most of the visual variation here is statistical noise, not
signal.

```{r p2-just-right, fig.height=3.5}
ggplot(pima, aes(x = pedigree)) +
  geom_histogram(binwidth = 0.1, fill = "#d97757", color = "white") +
  theme_minimal() +
  labs(title = "binwidth = 0.1 — just right",
       subtitle = "Right-skewed, mode around 0.25, long tail past 1.5",
       x = "Pedigree score", y = "Count")
```

Clean shape — mode, skew, and tail all visible.

```{r p2-too-coarse, fig.height=3.5}
ggplot(pima, aes(x = pedigree)) +
  geom_histogram(binwidth = 0.3, fill = "#9a958c", color = "white") +
  theme_minimal() +
  labs(title = "binwidth = 0.3 — too coarse",
       x = "Pedigree score", y = "Count")
```

The skew is hidden. The distribution looks almost symmetric, which is
misleading.

---

# Problem 3 — Diastolic pressure by group

```{r p3-counts}
pima |> count(diabetes)
```

n = 484 healthy, 268 diabetic. Both groups large, so a violin plot will
work better than `geom_jitter` (which would over-plot at this size).

```{r p3-chart}
ggplot(pima, aes(x = diabetes, y = pressure, fill = diabetes)) +
  geom_violin(alpha = 0.5, color = NA) +
  geom_boxplot(width = 0.15, fill = "white", outlier.shape = NA) +
  scale_fill_manual(values = c("neg" = "#9a958c", "pos" = "#d97757")) +
  scale_x_discrete(labels = c("neg" = "Healthy", "pos" = "Diabetic")) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    title    = "Diabetic patients have modestly higher diastolic pressure",
    subtitle = "Distributions overlap heavily — the difference is small",
    x        = NULL,
    y        = "Diastolic blood pressure (mmHg)"
  )
```

```{r p3-stats}
pima |>
  group_by(diabetes) |>
  summarise(
    n         = n(),
    median_bp = median(pressure),
    Q1        = quantile(pressure, 0.25),
    Q3        = quantile(pressure, 0.75),
    IQR       = IQR(pressure)
  )
```

Healthy: median 70 mmHg, IQR 64–78. Diabetic: median 74, IQR 68–82. The
diabetic group has slightly higher diastolic pressure on average, but the
distributions overlap substantially. Pressure alone is not a strong
discriminator.

---

# Problem 4 — Catch the lie

The lie is the truncated y-axis. It starts at 110 instead of 0.

```{r p4-numbers}
pima |>
  group_by(diabetes) |>
  summarise(mean_glucose = mean(glucose))
```

Numerically: 110.0 vs 142.3, a ratio of ~1.3. The diabetic mean is 30%
higher than the healthy mean, not 400% higher.

In the bad chart, the diabetic bar is roughly 4–5× the height of the
healthy bar. **Lie factor ≈ 4.5 / 1.3 ≈ 3.5.** The chart visually
exaggerates the difference by about three and a half times.

The fix: let y start at zero so the visual ratio matches the numerical
ratio.

```{r p4-fix}
group_summary <- pima |>
  group_by(diabetes) |>
  summarise(mean_glucose = mean(glucose))

ggplot(group_summary, aes(x = diabetes, y = mean_glucose, fill = diabetes)) +
  geom_col(width = 0.6) +
  scale_y_continuous(expand = expansion(mult = c(0, .05))) +
  scale_fill_manual(values = c("neg" = "#9a958c", "pos" = "#d97757")) +
  scale_x_discrete(labels = c("neg" = "Healthy", "pos" = "Diabetic")) +
  theme_minimal() +
  theme(legend.position    = "none",
        panel.grid.major.x = element_blank(),
        panel.grid.minor   = element_blank()) +
  labs(
    title    = "Diabetic patients average ~30% higher glucose",
    subtitle = "Mean fasting glucose, Pima Indians cohort",
    x        = NULL,
    y        = "Mean glucose (mg/dL)"
  )
```

Notice the difference is still clearly visible — it just isn't
sensationalized. The honest chart tells you what's true.

---

# Problem 5 — Three solutions, one per option

Any of the three is a fine choice. They each test something slightly
different.

## Option A — Age by group

```{r p5-age}
ggplot(pima, aes(x = diabetes, y = age, fill = diabetes)) +
  geom_violin(alpha = 0.5, color = NA) +
  geom_boxplot(width = 0.15, fill = "white", outlier.shape = NA) +
  scale_fill_manual(values = c("neg" = "#9a958c", "pos" = "#d97757")) +
  scale_x_discrete(labels = c("neg" = "Healthy", "pos" = "Diabetic")) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(
    title    = "Diabetic patients are older — median 36 vs 27",
    subtitle = "The healthy group skews young; the diabetic group is more uniform",
    x        = NULL,
    y        = "Age (years)"
  )
```

## Option B — Glucose × BMI by group

The two-by-two scatter is more informative than a single dimension because
it shows where the diagnosis lives in the joint space.

```{r p5-scatter}
ggplot(pima, aes(x = glucose, y = mass, color = diabetes)) +
  geom_point(alpha = 0.55, size = 1.8) +
  scale_color_manual(values = c("neg" = "#9a958c", "pos" = "#d97757"),
                     labels = c("Healthy", "Diabetic")) +
  theme_minimal() +
  theme(legend.position = "top") +
  labs(
    title    = "Diabetic patients cluster at high glucose AND high BMI",
    subtitle = "The diagnosis lives in the upper-right corner of the joint distribution",
    x        = "Glucose (mg/dL)",
    y        = "BMI (kg/m²)",
    color    = NULL
  )
```

## Option C — Pregnancies and diabetes

This one is a trap if you plot raw counts — the groups are unequal in
size, so raw counts are uninterpretable. The right framing is *proportion
diabetic* within each pregnancy group.

```{r p5-pregnancies}
pima |>
  mutate(
    pregnant_grp = case_when(
      pregnant == 0 ~ "0",
      pregnant <= 2 ~ "1–2",
      pregnant <= 5 ~ "3–5",
      TRUE          ~ "6+"
    ),
    pregnant_grp = factor(pregnant_grp, levels = c("0", "1–2", "3–5", "6+"))
  ) |>
  count(pregnant_grp, diabetes) |>
  group_by(pregnant_grp) |>
  mutate(prop = n / sum(n)) |>
  filter(diabetes == "pos") |>
  ggplot(aes(x = pregnant_grp, y = prop)) +
  geom_col(fill = "#d97757") +
  scale_y_continuous(labels = percent_format(),
                     expand = expansion(mult = c(0, .05))) +
  theme_minimal() +
  theme(panel.grid.major.x = element_blank()) +
  labs(
    title    = "Diabetes prevalence rises sharply with pregnancies",
    subtitle = "From ~16% in nulliparous patients to ~55% with 6 or more",
    x        = "Number of pregnancies",
    y        = "Proportion diabetic"
  )
```
