---
title: "Data Visualization — Practice Problems"
subtitle: "R for Clinicians · Lecture 07"
author: "Optional · for those who want to practice"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)
```

# About these problems

These are practice problems, not graded homework. Solve any of them you
find interesting. Skip any that don't speak to you.

Solutions are in a separate file (`visualization_solutions.Rmd`) — try the
problem first, then peek.

## Setup

```{r setup-data}
library(tidyverse)
library(mlbench)

data("PimaIndiansDiabetes")

# Same cleaning step from the workshop — zero-coded missing values removed
pima <- PimaIndiansDiabetes |>
  as_tibble() |>
  filter(glucose > 0, mass > 0, pressure > 0)

nrow(pima)
```

You will work with `pima` for all five problems. 752 patients, 9 variables.

---

# Problem 1 — A bar chart that earns its sort

Create a categorical variable `bmi_class` from the `mass` column using the
WHO BMI categories:

| Category        | Range          |
|-----------------|----------------|
| Underweight     | < 18.5         |
| Normal          | 18.5 – 24.9    |
| Overweight      | 25.0 – 29.9    |
| Obese class I   | 30.0 – 34.9    |
| Obese class II  | 35.0 – 39.9    |
| Obese class III | ≥ 40           |

Build a bar chart showing how many patients fall into each category.
Should you use `fct_reorder` here? Why or why not?

```{r p1}
# Your code:




```

---

# Problem 2 — Bin width as a decision

Build three histograms of `pedigree` (the diabetes pedigree function — a
score summarising family history). Use three different bin widths:

- `binwidth = 0.05`
- `binwidth = 0.1`
- `binwidth = 0.3`

Which one would you publish in a paper? What does each of the other two
hide or exaggerate?

```{r p2}
# Your code:




```

---

# Problem 3 — Comparing groups, honestly

Compare `pressure` (diastolic blood pressure, mmHg) between diabetic and
non-diabetic patients.

a) Build a boxplot.
b) Add either `geom_jitter` or switch to a violin plot — choose based on
   the sample size.
c) Compute the median and IQR for each group with `dplyr`.
d) The title of the chart should state the finding, not "boxplot of pressure".

```{r p3}
# Your code:




```

---

# Problem 4 — Catch the lie

Below is a chart someone built. It is technically correct — the numbers
are real — but it lies visually.

```{r p4-bad}
group_summary <- pima |>
  group_by(diabetes) |>
  summarise(mean_glucose = mean(glucose))

ggplot(group_summary, aes(x = diabetes, y = mean_glucose, fill = diabetes)) +
  geom_col() +
  coord_cartesian(ylim = c(110, 145)) +
  scale_fill_manual(values = c("neg" = "#c45e5e", "pos" = "#c45e5e")) +
  theme_gray() +
  labs(title = "Glucose by group", x = NULL, y = "Glucose")
```

What is the lie? Estimate the lie factor (visual ratio ÷ numerical ratio).
Then rebuild the chart honestly.

```{r p4-fix}
# Your code:




```

---

# Problem 5 — Your own finding

Pick **one** of the following. Build a single chart that makes the
finding obvious to a clinician with thirty seconds and no patience:

- Does `age` differ between diabetic and non-diabetic patients?
- Does the relationship between `glucose` and `mass` look different for
  diabetic versus non-diabetic patients?
- Are higher numbers of pregnancies associated with diabetes?

```{r p5}
# Your code:




```
