---
title: "The Grammar of Seeing — Workshop"
subtitle: "Companion code for the data visualization lecture"
author: "R for Clinicians · Lecture 07"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 2
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center",
  dpi = 100
)
```

# Before we start

This file is the workshop companion to the lecture deck. The deck makes the
argument; this file builds the charts. We use the same dataset throughout —
`PimaIndiansDiabetes` — so the only thing changing from chunk to chunk is the
visualization choice.

> **The rule of this workshop:** every chunk should change exactly one thing.
> Watch what happens.
>
> Homework is in a separate file (`visualization_homework.Rmd`).

## Packages

```{r packages}
library(tidyverse)   # ggplot2, dplyr, forcats, tibble
library(mlbench)     # PimaIndiansDiabetes dataset
library(scales)      # nicer axis formatting
```

If any of these are missing:

```{r install, eval=FALSE}
install.packages(c("tidyverse", "mlbench", "scales"))
```

## The data

```{r data}
data("PimaIndiansDiabetes")
pima <- as_tibble(PimaIndiansDiabetes)

pima
```

A quick orientation:

```{r data-glance}
glimpse(pima)
```

768 patients, 8 numeric variables, 1 categorical outcome (`diabetes`).
This is enough variety to demonstrate every chart type we care about.

---

# Part I — Why we visualize

## Anscombe's Quartet — same statistics, different stories

Before we touch `pima`, a quick demonstration of why summary statistics aren't
enough. R ships with Anscombe's Quartet built in.

```{r anscombe-stats}
anscombe_long <- anscombe |>
  pivot_longer(everything(),
               names_to  = c(".value", "set"),
               names_pattern = "(.)(.)")

# Verify all four datasets have nearly identical statistics
anscombe_long |>
  group_by(set) |>
  summarise(
    mean_x = mean(x),
    mean_y = round(mean(y), 2),
    sd_x   = sd(x),
    sd_y   = round(sd(y), 2),
    cor    = round(cor(x, y), 3)
  )
```

Now plot them:

```{r anscombe-plot}
ggplot(anscombe_long, aes(x, y)) +
  geom_point(size = 2.5, color = "#d97757") +
  geom_smooth(method = "lm", se = FALSE, color = "#4a4640") +
  facet_wrap(~ set, labeller = label_both) +
  theme_minimal()
```

**Identical mean, sd, correlation, regression line. Four completely different
worlds.** This is why we don't trust the table alone.

---

# Part II — Building in ggplot2

## The grammar in one chart

Every ggplot is built in layers:

```{r grammar-explicit}
ggplot(data = pima) +                            # 1. data
  aes(x = glucose, y = mass) +                   # 2. aesthetics
  geom_point(alpha = 0.4, color = "#d97757") +   # 3. geometry
  theme_minimal() +                              # 4. refinement
  labs(x = "Glucose (mg/dL)", y = "BMI (kg/m²)")
```

## Why continuous goes to position, categorical goes to colour

```{r mapping-logic}
ggplot(pima, aes(x = glucose, y = mass, color = diabetes)) +
  geom_point(alpha = 0.5) +
  scale_color_manual(values = c("neg" = "#9a958c", "pos" = "#d97757")) +
  theme_minimal() +
  labs(
    x = "Glucose (mg/dL)",
    y = "BMI (kg/m²)",
    color = "Diabetes"
  )
```

Position carries the two continuous measurements (the precise channels).
Colour carries the categorical group (the weaker channel, but enough to
distinguish two classes).

---

# Part III — The three winning charts

## 1. Bar chart — comparing categories

We need a categorical variable to count. The Pima dataset only has `diabetes`,
so let's build a more interesting one by binning age:

```{r bar-data}
pima_age <- pima |>
  mutate(
    age_group = cut(age,
                    breaks = c(20, 30, 40, 50, 60, 90),
                    labels = c("21–30", "31–40", "41–50", "51–60", "61+"),
                    right  = FALSE)
  )

age_counts <- pima_age |>
  count(age_group)

age_counts
```

### Naive bar chart

```{r bar-v1}
ggplot(age_counts, aes(x = age_group, y = n)) +
  geom_col()
```

### Apply the three rules

**Rule 1 — y starts at zero (already does for `geom_col`, but be explicit):**

```{r bar-v2}
ggplot(age_counts, aes(x = age_group, y = n)) +
  geom_col(fill = "#d97757") +
  scale_y_continuous(expand = expansion(mult = c(0, .05)))
```

**Rule 2 — sort by value with `fct_reorder`:**

```{r bar-v3}
ggplot(age_counts, aes(x = fct_reorder(age_group, n), y = n)) +
  geom_col(fill = "#d97757") +
  scale_y_continuous(expand = expansion(mult = c(0, .05)))
```

> **When NOT to sort:** ordinal categories like age groups have a natural order.
> For a real ordered category we'd skip `fct_reorder`. Switch to a different
> example to see sorting earn its keep:

```{r bar-sort-correct}
# Imagine departments instead of ages — no inherent order
dept <- tibble(
  department = c("Cardiology", "Internal Med A", "Emergency",
                 "Neurology", "Paediatric ICU", "Oncology"),
  n = c(142, 198, 87, 64, 41, 113)
)

ggplot(dept, aes(x = fct_reorder(department, n), y = n)) +
  geom_col(fill = "#d97757") +
  scale_y_continuous(expand = expansion(mult = c(0, .05)))
```

**Rule 3 — long labels? `coord_flip()`:**

```{r bar-v4}
ggplot(dept, aes(x = fct_reorder(department, n), y = n)) +
  geom_col(fill = "#d97757") +
  coord_flip() +
  scale_y_continuous(expand = expansion(mult = c(0, .05))) +
  labs(x = NULL, y = "Patients") +
  theme_minimal()
```

---

## 2. Histogram — distribution shape

### Default — what does ggplot2 give us?

```{r hist-v1}
ggplot(pima, aes(x = glucose)) +
  geom_histogram()
```

ggplot warns: *"stat_bin() using `bins = 30`. Pick better value with
`binwidth`."* It's right. 30 is arbitrary.

### Filter the zeros

Glucose = 0 is impossible — those are missing values coded as zero.
Same for `mass`, `pressure`, `triceps`, `insulin`. Always check.

```{r hist-clean}
pima_clean <- pima |>
  filter(glucose > 0, mass > 0, pressure > 0)

nrow(pima)         # before
nrow(pima_clean)   # after
```

### Bin width matters

```{r hist-binwidth-coarse, fig.height=3.5}
ggplot(pima_clean, aes(x = glucose)) +
  geom_histogram(binwidth = 30, fill = "#c45e5e", color = "white")
```

Too coarse — the structure is hidden.

```{r hist-binwidth-fine, fig.height=3.5}
ggplot(pima_clean, aes(x = glucose)) +
  geom_histogram(binwidth = 2, fill = "#c45e5e", color = "white")
```

Too fine — pure noise.

```{r hist-binwidth-good, fig.height=3.5}
ggplot(pima_clean, aes(x = glucose)) +
  geom_histogram(binwidth = 5, fill = "#d97757", color = "white") +
  theme_minimal()
```

Right. Now we can see the shape: roughly normal around 110, with a long
right tail and a small bump near 100. That bump is interesting — it's the
diabetic group leaking into the healthy distribution.

### Why `color = "white"` matters

The white outline acts as a separator between bins. It costs no data ink
but gains huge readability.

---

## 3. Box plot — beyond the average

### Single boxplot

```{r box-v1}
ggplot(pima_clean, aes(y = glucose)) +
  geom_boxplot()
```

That's the five-number summary in one shape: min, Q1, median, Q3, max,
plus outliers as dots.

### Compared across groups — this is where boxplots earn their keep

```{r box-v2}
ggplot(pima_clean, aes(x = diabetes, y = glucose, fill = diabetes)) +
  geom_boxplot(alpha = 0.6, outlier.color = "#c45e5e") +
  scale_fill_manual(values = c("neg" = "#9a958c", "pos" = "#d97757")) +
  theme_minimal() +
  labs(x = NULL, y = "Glucose (mg/dL)") +
  theme(legend.position = "none")
```

Diabetic patients have higher glucose. The medians don't even overlap with
the boxes of the other group — that's a strong visual signal.

### Upgrade 1: when there are few points, add `geom_jitter`

```{r box-jitter}
pima_small <- pima_clean |>
  group_by(diabetes) |>
  slice_sample(n = 30) |>
  ungroup()

ggplot(pima_small, aes(x = diabetes, y = glucose, fill = diabetes)) +
  geom_boxplot(alpha = 0.4, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.7, size = 2) +
  scale_fill_manual(values = c("neg" = "#9a958c", "pos" = "#d97757")) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(x = NULL, y = "Glucose (mg/dL)",
       subtitle = "n = 30 per group, individual points shown")
```

### Upgrade 2: when there are many points, switch to violin

```{r violin}
ggplot(pima_clean, aes(x = diabetes, y = glucose, fill = diabetes)) +
  geom_violin(alpha = 0.5, color = NA) +
  geom_boxplot(width = 0.15, fill = "white", outlier.shape = NA) +
  scale_fill_manual(values = c("neg" = "#9a958c", "pos" = "#d97757")) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(x = NULL, y = "Glucose (mg/dL)")
```

The violin reveals what the box hides — the shape of the distribution
inside the IQR. Notice the diabetic group's density is bimodal-ish; the
boxplot alone would never show that.

---

# Part IV — Refining and storytelling

We'll take one chart and walk it through a refinement pipeline.

## v1 — The default

```{r refine-v1}
ggplot(pima_clean, aes(x = glucose, fill = diabetes)) +
  geom_histogram(binwidth = 5)
```

## v2 — Theme and palette

```{r refine-v2}
ggplot(pima_clean, aes(x = glucose, fill = diabetes)) +
  geom_histogram(binwidth = 5, color = "white", alpha = 0.8) +
  scale_fill_manual(values = c("neg" = "#9a958c", "pos" = "#d97757")) +
  theme_minimal()
```

## v3 — Title states the finding

```{r refine-v3}
ggplot(pima_clean, aes(x = glucose, fill = diabetes)) +
  geom_histogram(binwidth = 5, color = "white", alpha = 0.8) +
  scale_fill_manual(values = c("neg" = "#9a958c", "pos" = "#d97757")) +
  labs(
    title    = "Diabetic patients cluster at higher glucose levels",
    subtitle = "Pima Indians cohort · n = 752 after removing zero-coded missing values",
    x        = "Glucose (mg/dL)",
    y        = "Number of patients",
    fill     = "Diabetes",
    caption  = "Source: PimaIndiansDiabetes, mlbench package"
  ) +
  theme_minimal()
```

## v4 — Strategic colour: highlight one group only

```{r refine-v4}
ggplot(pima_clean, aes(x = glucose, fill = diabetes)) +
  geom_histogram(binwidth = 5, color = "white", alpha = 0.85,
                 position = "identity") +
  scale_fill_manual(
    values = c("neg" = "#cdc8bf", "pos" = "#d97757"),
    labels = c("Healthy", "Diabetic")
  ) +
  labs(
    title    = "Diabetic patients cluster at higher glucose levels",
    subtitle = "Pima Indians cohort · n = 752",
    x        = "Glucose (mg/dL)",
    y        = "Number of patients",
    fill     = NULL,
    caption  = "Source: PimaIndiansDiabetes, mlbench"
  ) +
  theme_minimal() +
  theme(
    legend.position    = "top",
    plot.title         = element_text(face = "bold", size = 14),
    plot.subtitle      = element_text(color = "grey40"),
    panel.grid.minor   = element_blank(),
    panel.grid.major.x = element_blank()
  )
```

The diabetic distribution is the message — colour it warmly. The healthy
distribution is the comparison baseline — let it recede.

## Viridis for continuous scales

When you genuinely need a colour gradient (heatmaps, density plots,
continuous variables encoded as colour), use viridis. It's perceptually
uniform and colour-blind safe.

```{r viridis}
ggplot(pima_clean, aes(x = glucose, y = mass, color = age)) +
  geom_point(alpha = 0.7, size = 2) +
  scale_color_viridis_c(option = "magma") +
  theme_minimal() +
  labs(
    x = "Glucose (mg/dL)",
    y = "BMI (kg/m²)",
    color = "Age",
    title = "Older patients drift toward the high-glucose, high-BMI corner"
  )
```

---

# The lie factor — what NOT to do

This chunk is intentionally bad. Don't ship this.

```{r lie-factor, fig.height=4}
# Truncated y-axis — exaggerates differences
group_means <- pima_clean |>
  group_by(diabetes) |>
  summarise(mean_glucose = mean(glucose))

ggplot(group_means, aes(x = diabetes, y = mean_glucose, fill = diabetes)) +
  geom_col() +
  coord_cartesian(ylim = c(100, 145)) +   # <-- THE LIE
  scale_fill_manual(values = c("neg" = "#c45e5e", "pos" = "#c45e5e")) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "DON'T DO THIS — y axis truncated",
       x = NULL, y = "Glucose")
```

Compare to the honest version:

```{r honest, fig.height=4}
ggplot(group_means, aes(x = diabetes, y = mean_glucose, fill = diabetes)) +
  geom_col() +
  scale_y_continuous(expand = expansion(mult = c(0, .05))) +
  scale_fill_manual(values = c("neg" = "#9a958c", "pos" = "#6cb585")) +
  theme_minimal() +
  theme(legend.position = "none") +
  labs(title = "Honest version — y starts at zero",
       x = NULL, y = "Mean glucose (mg/dL)")
```

Same data. Same comparison. The first chart screams; the second informs.
The lie factor of the first is roughly 4 — the visual difference is four
times larger than the numerical difference.

---

# Cheat sheet — the rules in one place

```{r cheatsheet, echo=FALSE}
tribble(
  ~Rule, ~Code,
  "y axis starts at zero",     'scale_y_continuous(expand = expansion(mult = c(0, .05)))',
  "Sort bars by value",        'fct_reorder(category, value)',
  "Long labels? Flip",         'coord_flip()',
  "Bin width with units",      'geom_histogram(binwidth = 5, color = "white")',
  "See individual points",     'geom_jitter(width = 0.15)',
  "See full distribution",     'geom_violin() + geom_boxplot(width = 0.15)',
  "Title = the finding",       'labs(title = "Diabetic patients cluster higher")',
  "Units in axis labels",      'labs(x = "Glucose (mg/dL)")',
  "Highlight, don\'t decorate", 'scale_fill_manual(values = c("focus" = "#d97757", "rest" = "grey80"))',
  "Continuous colour scale",   'scale_color_viridis_c()',
  "Strip the gridlines",       'theme_minimal() + theme(panel.grid.minor = element_blank())'
)
```

---

# Session info

For reproducibility:

```{r session}
sessionInfo()
```
