---
title: "Evaluating Models & Tuning Them"
subtitle: "R for Clinicians — live session"
output:
  html_document:
    toc: true
    toc_float: true
    theme: darkly
    highlight: zenburn
    df_print: kable
---

```{r setup, include=FALSE}
# Instructor walk-through: every chunk runs top-to-bottom during class.
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 6, fig.height = 4.5, fig.align = "center")

# palette matching the slides, for consistent plots
col_green <- "#4ade80"; col_cyan <- "#38bdf8"
col_amber <- "#fbbf24"; col_red  <- "#f87171"
```

# Setup

```{r libs}
library(mlbench)   # PimaIndiansDiabetes
library(caret)     # train(), confusionMatrix(), resamples()
library(pROC)      # roc(), auc(), coords()

data("PimaIndiansDiabetes")
d <- PimaIndiansDiabetes

str(d)
table(d$diabetes)          # pos / neg  — our two classes
prop.table(table(d$diabetes))
```

`pos` (diabetes present) is our **positive class** all session. A false
negative = a diabetic patient told they're fine.

---

# 1 · Split — the lab bench  *(slide 4)*

The test set is opened **once**, at the very end. It stands in for tomorrow's
patients.

```{r split}
set.seed(42)
idx   <- createDataPartition(d$diabetes, p = 0.7, list = FALSE)
train <- d[idx, ]
test  <- d[-idx, ]

nrow(train); nrow(test)
table(test$diabetes)       # our test cohort
```

`createDataPartition` stratifies — the pos/neg ratio is preserved in both
sets.

---

# 2 · The accuracy trap  *(slides 5–9)*

## Model A — "everyone's healthy"

```{r modelA}
# predict the majority class for every patient
modelA_pred <- factor(rep("neg", nrow(test)), levels = levels(test$diabetes))

confusionMatrix(modelA_pred, test$diabetes, positive = "pos")
```

Notice the accuracy — it looks respectable, purely because most patients are
`neg`. Now read the **Sensitivity** line: it is 0. Model A catches no
diabetics at all. *This is why accuracy alone lies.*

## Model B — our decision tree

```{r modelB}
fit_tree <- train(diabetes ~ ., data = train, method = "rpart")

pred_B <- predict(fit_tree, newdata = test)
cm_B   <- confusionMatrix(pred_B, test$diabetes, positive = "pos")
cm_B
```

Model B's accuracy may be **lower** than Model A's — yet it's the only one
that catches any disease. Accuracy is the wrong headline.

---

# 3 · The four numbers  *(slides 10–11)*

Everything comes from the 2×2 table. Let's pull it apart by hand, then check
against caret.

```{r metrics-by-hand}
tab <- cm_B$table
TP <- tab["pos", "pos"]; FN <- tab["neg", "pos"]
FP <- tab["pos", "neg"]; TN <- tab["neg", "neg"]
c(TP = TP, FP = FP, FN = FN, TN = TN)

sensitivity <- TP / (TP + FN)   # of the truly sick, how many caught
specificity <- TN / (TN + FP)   # of the truly healthy, how many cleared
ppv         <- TP / (TP + FP)   # when it says +, how often right
c(sensitivity = sensitivity, specificity = specificity, PPV = ppv)

# caret computed the same thing:
cm_B$byClass[c("Sensitivity", "Specificity", "Pos Pred Value")]
```

---

# 4 · The threshold is a choice  *(slides 12–13)*

The tree doesn't really output "pos/neg" — it outputs a **probability**, and
a cutoff turns it into a decision. The default cutoff is 0.5.

```{r probs}
prob <- predict(fit_tree, newdata = test, type = "prob")[, "pos"]
head(prob)
```

Watch what happens to the confusion matrix as we move the line. Same model,
same probabilities — different decisions.

```{r threshold-sweep}
make_matrix <- function(p, cutoff) {
  pred <- factor(ifelse(p > cutoff, "pos", "neg"),
                 levels = levels(test$diabetes))
  cm <- confusionMatrix(pred, test$diabetes, positive = "pos")
  round(cm$byClass[c("Sensitivity", "Specificity")], 3)
}

rbind(
  `cutoff 0.5` = make_matrix(prob, 0.5),
  `cutoff 0.3` = make_matrix(prob, 0.3),
  `cutoff 0.7` = make_matrix(prob, 0.7)
)
```

Drop the cutoff → catch more disease (sensitivity ↑), more false alarms
(specificity ↓). Raise it → the reverse. **Which line?** is a clinical
question — hold it.

---

# 5 · ROC & AUC  *(slides 14–16)*

The ROC curve is every cutoff at once. Each point is one confusion matrix.

```{r roc}
roc_obj <- roc(response  = test$diabetes,
               predictor = prob,
               levels    = c("neg", "pos"))

plot(roc_obj, print.auc = TRUE, col = col_green, lwd = 3,
     legacy.axes = TRUE)     # x = 1 - specificity

auc(roc_obj)
```

The two cutoffs we computed by hand (0.5 and 0.3) are two points on this
curve; the line fills in all the others. **AUC** = the whole curve as one
number — the probability the model ranks a random `pos` above a random `neg`.
Threshold-free, prevalence-free.

---

# 6 · So — which threshold?  *(slide 17)*

Youden's J = sensitivity + specificity − 1: the point farthest from the
diagonal. A reasonable default — *not* an answer.

```{r youden}
coords(roc_obj, x = "best", best.method = "youden",
       ret = c("threshold", "sensitivity", "specificity"))
```

This assumes a missed diabetic and a false alarm cost the same. They don't.
The clinic sets the line; statistics only proposes.

---

# 7 · Don't trust one split — cross-validation  *(slides 20–21)*

Was that AUC skill, or a lucky split? k-fold CV rotates the held-out fold and
averages, so no single split can fool us. This also becomes the **engine of
tuning**.

```{r cv-control}
ctrl <- trainControl(method          = "cv",
                     number           = 10,
                     classProbs       = TRUE,        # keep probabilities
                     summaryFunction  = twoClassSummary)  # report AUC

fit_tree_cv <- train(diabetes ~ ., data = train,
                     method    = "rpart",
                     trControl = ctrl,
                     metric    = "ROC")   # optimize AUC, not accuracy

fit_tree_cv$results
```

`metric = "ROC"` — the whole first hour in one argument: stop optimizing
accuracy, optimize AUC.

---

# 8 · Parameters vs hyperparameters  *(slides 24–25)*

Parameters are **learned** from data (the tree's split points). Hyperparameters
are **chosen** before training (`cp`, the complexity parameter). Tuning =
searching for the hyperparameter values that maximize cross-validated AUC.

```{r tune-cp}
grid_cp <- expand.grid(cp = seq(0.001, 0.10, by = 0.005))

fit_tuned <- train(diabetes ~ ., data = train,
                   method    = "rpart",
                   trControl = ctrl,
                   metric    = "ROC",
                   tuneGrid  = grid_cp)

fit_tuned$bestTune
plot(fit_tuned)   # the tuning curve — the U-curve in the wild
```

The peak is `bestTune`. Left of it: underfit. Right: overfit. Same data —
better dial.

```{r tune-gain}
c(default_cp_AUC = max(fit_tree_cv$results$ROC),
  tuned_AUC      = max(fit_tuned$results$ROC))
```

---

# 9 · One interface, any model  *(slides 29–30)*

The caret contract: change `method` and the grid, keep everything else. Note
kNN needs `center`/`scale` — and caret applies it **inside each fold**, which
prevents preprocessing leakage.

```{r other-models}
fit_knn <- train(diabetes ~ ., data = train,
                 method     = "knn",
                 preProcess = c("center", "scale"),
                 trControl  = ctrl, metric = "ROC",
                 tuneGrid   = expand.grid(k = seq(3, 31, by = 2)))

fit_rf <- train(diabetes ~ ., data = train,
                method    = "rf",
                trControl = ctrl, metric = "ROC",
                tuneGrid  = expand.grid(mtry = 2:6))

fit_knn$bestTune; fit_rf$bestTune
```

## Fair comparison — same folds, same metric

```{r resamples}
res <- resamples(list(tree = fit_tuned, knn = fit_knn, rf = fit_rf))
summary(res)
bwplot(res, metric = "ROC")   # look at the spread, not just the means
```

The winner is the highest mean ROC — but if the boxplots overlap, the gap may
be noise, and a simpler model is a fair choice.

---

# 10 · The final grade — test set, once  *(slides 31–33)*

Everything above happened **inside the training data** via CV. Only now do we
open the test set — once — with the model CV chose.

```{r final}
cv_auc <- c(tree = max(fit_tuned$results$ROC),
            knn  = max(fit_knn$results$ROC),
            rf   = max(fit_rf$results$ROC))
best_name  <- names(which.max(cv_auc))
best_model <- list(tree = fit_tuned, knn = fit_knn, rf = fit_rf)[[best_name]]
best_name

# ---- test set opened ONCE ----
final_pred <- predict(best_model, newdata = test)
confusionMatrix(final_pred, test$diabetes, positive = "pos")

final_prob <- predict(best_model, newdata = test, type = "prob")[, "pos"]
final_roc  <- roc(test$diabetes, final_prob, levels = c("neg", "pos"))
c(test_AUC = as.numeric(auc(final_roc)),
  cv_AUC   = max(best_model$results$ROC))
```

If test AUC and CV AUC are close, the estimate is trustworthy. A large drop
would signal overfitting or leakage — the model wouldn't reproduce in the
next hospital.

---

# The pipeline, in one place  *(slide 33)*

> **Split → CV + tune → compare → final fit → test once → choose threshold.**

Every honest clinical prediction model can draw this diagram of itself. The
test set answers exactly one question, exactly once: *how will this model do
on patients it has never seen?*

```{r session-info}
sessionInfo()
```
