---
title: "Practice Lab — Evaluating Models & Tuning Them"
subtitle: "R for Clinicians · self-paced, not submitted"
output:
  html_document:
    toc: true
    toc_float: true
    theme: darkly
    highlight: zenburn
---

```{r setup, include=FALSE}
# eval = FALSE so this file knits before you write your code.
# Change to eval = TRUE (or remove) once your code is ready.
knitr::opts_chunk$set(eval = FALSE, message = FALSE, warning = FALSE)
```

# How to use this lab

Nothing here is submitted and nothing is graded. This is the lecture pipeline,
run by your own hands on a dataset you haven't seen — that is the only way the
ideas stick.

Work one exercise at a time. When you're stuck for more than a few minutes,
open the walkthrough (`class14_solution_evaluation_tuning.Rmd`) for that
exercise only, then come back. Reading the walkthrough top to bottom teaches you
nothing; you'll recognize every line and be able to write none of them.

Each exercise ends with a **Check yourself** box. If your answer to that box is
"I don't know", that's the signal to go back to the slides — not to move on.

## Packages

Install once, before you start. `caret` doesn't bundle the modeling engines —
it calls them — so the engines need to be present too:

```{r install, eval=FALSE}
install.packages(c(
  "mlbench",       # the datasets
  "caret",         # the pipeline
  "pROC",          # ROC / AUC
  "rpart",         # engine for method = "rpart"
  "randomForest",  # engine for method = "rf"
  "e1071",         # caret dependency for confusionMatrix
  "kknn"           # engine some knn variants need
))
```

If `randomForest` fails to compile, use
`install.packages("randomForest", type = "binary")` on Windows/macOS.

## The data

`BreastCancer` — benign vs malignant tumors from cytology features. The
cleaning is done for you; run it as-is:

```{r data-prep, eval=TRUE}
library(mlbench)
library(caret)
library(pROC)

data(BreastCancer)
bc <- BreastCancer

bc$Id <- NULL                                   # Exercise 15 asks why!
bc[, 1:9] <- lapply(bc[, 1:9],
                    function(x) as.numeric(as.character(x)))
bc <- na.omit(bc)

str(bc)          # 9 numeric predictors + Class (benign / malignant)
table(bc$Class)
```

Throughout: the **positive class is `"malignant"`** — a false negative means
telling a patient with cancer that everything is fine.

---

# Part 1 · Split

**1.** Split `bc` into 70% training and 30% test using
`createDataPartition()`. Set `set.seed(2026)` first so your results are
reproducible.

```{r ex1}
# your code here
```

> **Check yourself:** how many patients in the test set, how many malignant?
> Is the benign/malignant ratio the same as in the full data — and did you
> have to do anything to make that happen?

---

# Part 2 · A first model and its confusion matrix

**2.** Train a decision tree (`method = "rpart"`) on the training set with
default settings, predict classes on the test set, and print the confusion
matrix with `confusionMatrix(..., positive = "malignant")`.

```{r ex2}
# your code here
```

**3.** From the printed table, say TP, FP, FN, TN out loud in plain words
("X patients with cancer were caught", "Y healthy patients got a false
alarm", ...).

**4.** Compute sensitivity and specificity **by hand** from those four
numbers and check that they match caret's output.

```{r ex4}
# your code here
```

**5.** In *this* clinical setting, which red cell hurts more — FN or FP?

> **Check yourself:** if your hand-computed sensitivity doesn't match caret's,
> the usual cause is reading the confusion matrix table in the wrong
> orientation. Which margin is truth and which is prediction?

---

# Part 3 · The accuracy trap

**6.** "Model A" predicts **benign for every patient**. Compute Model A's
accuracy on your test set. Compute its sensitivity.

```{r ex6}
# your code here (hint: table(test$Class) is all you need)
```

> **Check yourself:** Model A's accuracy is high enough to put in a slide.
> Can you say in one sentence why that number is a lie?

---

# Part 4 · Thresholds, ROC, AUC

**7.** Get predicted **probabilities** of malignancy on the test set
(`type = "prob"`), build a ROC object with `pROC::roc()`, plot it, and
look at the AUC.

```{r ex7}
# your code here
```

**8.** Lower the classification threshold from 0.5 to **0.2** and rebuild
the confusion matrix.

```{r ex8}
# hint:
# pred_02 <- factor(ifelse(prob > 0.2, "malignant", "benign"),
#                   levels = levels(test$Class))
```

> **Check yourself:** sensitivity moved. Which direction, and what paid for it?
> Name the price in patients, not in percentage points.

**9.** Use `coords(roc_obj, "best")` to find the Youden-optimal threshold and
its sensitivity/specificity.

```{r ex9}
# your code here
```

> **Check yourself:** Youden treats a missed cancer and a false alarm as
> equally bad. Give one clinical reason you might deliberately **not** use the
> Youden threshold in a screening program.

---

# Part 5 · Cross-validation and tuning

**10.** Build a `trainControl` object for **10-fold cross-validation** that
keeps class probabilities and reports AUC (`twoClassSummary`). Re-train the
tree with `metric = "ROC"` and a `tuneGrid` for `cp` covering
`seq(0.0, 0.05, by = 0.005)`. Plot the tuning curve.

```{r ex10}
# your code here
```

> **Check yourself:** what does `metric = "ROC"` change about *which* model
> caret hands back? Does your tuning curve look like the U-curve from the
> slides — and if it doesn't, is that a bug or a fact about this dataset?

**11.** Tune a **kNN** model over odd values of k from 3 to 31.
kNN needs one extra argument the tree didn't.

```{r ex11}
# your code here
```

> **Check yourself:** which argument, and why does it matter that caret applies
> it *inside* each fold rather than once on the whole dataset?

**12.** Tune a **random forest** (`method = "rf"`) over `mtry = 2:6`.

```{r ex12}
# your code here
```

> **Check yourself:** if this errors with "Required packages are missing:
> randomForest", you skipped the install chunk. `caret` never bundles engines.

**13.** Compare the three tuned models with `resamples()`.

```{r ex13}
# your code here
```

> **Check yourself:** one model has the highest mean CV AUC. Now look at the
> boxplot spread — is the gap between the best two larger than the
> fold-to-fold noise? If not, what does "the winner" actually mean?

---

# Part 6 · Opening the test set

**14.** Take the winning model, predict **once** on the test set, and look at
the final confusion matrix and test AUC. Compare test AUC to CV AUC.

```{r ex14}
# your code here
```

> **Check yourself:** are they close? What would a large gap suggest — and in
> which direction does the gap usually go?

**15.** Two conceptual questions. Write your answers down before you look
anything up; the writing is the exercise.

a. In the data-prep chunk we deleted the `Id` column before modeling.
Suppose we had kept it and the model used it. Which lecture concept does
this violate, and why would the model look great in your report and fail in
the next hospital?

b. A colleague tunes five models, checks each one's **test-set** AUC, and
publishes the best. What rule was broken, and what is wrong with the
published number?

---

# If you only remember three things

1. Accuracy is a headline number that hides the errors that matter. Read the
   confusion matrix.
2. The threshold is a clinical decision, not a statistical default. 0.5 is not
   sacred.
3. Tune on cross-validation. The test set is opened once, at the end, and never
   again.
