---
title: "Walkthrough — Evaluating Models & Tuning Them"
subtitle: "R for Clinicians · companion to the practice lab"
output:
  html_document:
    toc: true
    toc_float: true
    theme: darkly
    highlight: zenburn
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(eval = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 6, fig.height = 4.5)
```

# How to use this

This is not an answer key to be marked against — nothing in the lab is
submitted or graded. It's a worked walkthrough for when you're stuck, and a
reference for the reasoning behind each step.

Read one exercise at a time, after you've tried it. Your numbers may differ
slightly from mine (different R version, different `rpart` build) — that's fine.
What should match is the *shape* of the answer and the reasoning.

## Packages

`caret` calls modeling engines rather than bundling them, so the engines must
be installed separately — this is the source of the
`Required packages are missing: randomForest` error at Exercise 12.

```{r install, eval=FALSE}
install.packages(c("mlbench", "caret", "pROC", "rpart",
                   "randomForest", "e1071", "kknn"))
```

# Data prep

```{r data-prep}
library(mlbench)
library(caret)
library(pROC)
library(randomForest)   # engine for method = "rf"

data(BreastCancer)
bc <- BreastCancer

bc$Id <- NULL
bc[, 1:9] <- lapply(bc[, 1:9],
                    function(x) as.numeric(as.character(x)))
bc <- na.omit(bc)

str(bc)
table(bc$Class)
```

`malignant` is the positive class throughout — a false negative is a missed
cancer.

---

# Part 1 · Split

**1.**

```{r ex1}
set.seed(2026)
idx   <- createDataPartition(bc$Class, p = 0.7, list = FALSE)
train <- bc[idx, ]
test  <- bc[-idx, ]

nrow(test)
table(test$Class)
```

`createDataPartition` splits **within each class**, so the benign/malignant
ratio in train and test matches the full data (stratified split) — important
when one class is rarer. You didn't have to ask for stratification; that's the
function's default behaviour on a factor, and it's the reason we don't use
`sample()` here.

---

# Part 2 · A first model and its confusion matrix

**2.**

```{r ex2}
fit_tree <- train(Class ~ ., data = train, method = "rpart")

pred <- predict(fit_tree, newdata = test)
cm   <- confusionMatrix(pred, test$Class, positive = "malignant")
cm
```

**3.** Reading the table (`cm$table`): the diagonal is correct, the
off-diagonal is errors. In words: TP = malignant tumors correctly flagged;
FN = malignant tumors called benign (missed cancers); FP = benign tumors
called malignant (false alarms → unnecessary biopsy/anxiety); TN = benign
tumors correctly cleared.

**4.**

```{r ex4}
tab <- cm$table
TP <- tab["malignant", "malignant"]
FN <- tab["benign",    "malignant"]
FP <- tab["malignant", "benign"]
TN <- tab["benign",    "benign"]

sensitivity_manual <- TP / (TP + FN)
specificity_manual <- TN / (TN + FP)

c(manual_sens = sensitivity_manual, caret_sens = cm$byClass["Sensitivity"])
c(manual_spec = specificity_manual, caret_spec = cm$byClass["Specificity"])
```

They match — caret is just automating the arithmetic we did on the slides.

**Check-yourself answer:** in `confusionMatrix(pred, truth)` the **rows are
predictions** and the **columns are truth**. Swap them and sensitivity and
specificity trade places — a mistake that survives peer review more often than
it should.

**5.** FN hurts more here: a missed malignancy delays treatment of a
life-threatening disease, while an FP leads to a follow-up biopsy —
distressing and costly, but recoverable. So we will prefer higher
sensitivity, even at some cost to specificity. *(The metric you optimize is a
clinical decision, not a statistical default.)*

---

# Part 3 · The accuracy trap

**6.**

```{r ex6}
n_test        <- nrow(test)
n_benign_test <- sum(test$Class == "benign")

modelA_accuracy    <- n_benign_test / n_test   # predicting benign for all
modelA_sensitivity <- 0                        # catches zero malignancies

c(accuracy = modelA_accuracy, sensitivity = modelA_sensitivity)
```

Model A scores a high accuracy purely because benign tumors are the majority
class — yet it catches **zero** cancers (sensitivity 0). Accuracy rewards the
majority class and hides total failure on the cases that matter. *(This is
slide 5–9 on your own data.)*

---

# Part 4 · Thresholds, ROC, AUC

**7.**

```{r ex7}
prob    <- predict(fit_tree, newdata = test, type = "prob")[, "malignant"]
roc_obj <- roc(response = test$Class, predictor = prob,
               levels = c("benign", "malignant"))

auc(roc_obj)
plot(roc_obj, print.auc = TRUE, col = "#4ade80")
```

`type = "prob"` is the switch — the ROC curve is built by sweeping a
threshold over these probabilities.

**8.**

```{r ex8}
pred_02 <- factor(ifelse(prob > 0.2, "malignant", "benign"),
                  levels = levels(test$Class))
confusionMatrix(pred_02, test$Class, positive = "malignant")$byClass[
  c("Sensitivity", "Specificity")]
```

Lowering the threshold catches more malignancies (**sensitivity ↑**) but
flags more benign tumors as suspicious (**specificity ↓**) — the same
trade-off Patient #47 illustrated on the slides, in the opposite direction.

**Check-yourself answer:** state the price as patients, not percentages —
"we now send *n* more women with benign tumors for a biopsy in order to catch
*m* more cancers". That sentence is the one you'd have to defend in a
tumor board; the percentage is the one that hides behind a decimal point.

**9.**

```{r ex9}
coords(roc_obj, x = "best", best.method = "youden",
       ret = c("threshold", "sensitivity", "specificity"))
```

Youden maximizes sensitivity + specificity − 1, treating a missed cancer and
a false alarm as equally bad. In cancer screening they are **not** equal —
we may deliberately choose a lower threshold than Youden to push sensitivity
higher, accepting more false alarms to miss fewer cancers.

---

# Part 5 · Cross-validation and tuning

**10.**

```{r ex10}
ctrl <- trainControl(method = "cv", number = 10,
                     classProbs = TRUE,
                     summaryFunction = twoClassSummary)

grid_cp <- expand.grid(cp = seq(0.0, 0.05, by = 0.005))

fit_tree_cv <- train(Class ~ ., data = train,
                     method    = "rpart",
                     trControl = ctrl,
                     metric    = "ROC",
                     tuneGrid  = grid_cp)

fit_tree_cv$bestTune
plot(fit_tree_cv)
```

`metric = "ROC"` optimizes AUC, not accuracy — the whole first hour, in one
argument. It changes which row of `fit_tree_cv$results` gets returned as
`bestTune`.

**Check-yourself answer:** the tuning curve here may *not* show a clean U.
That's not a bug — `BreastCancer` separates so easily that AUC is flat and high
across most of the `cp` range. A flat curve is itself information: the
hyperparameter doesn't matter much on this problem, so pick the simplest model
in the flat region rather than the arithmetic maximum.

**11.**

```{r ex11}
fit_knn <- train(Class ~ ., data = train,
                 method     = "knn",
                 preProcess = c("center", "scale"),
                 trControl  = ctrl,
                 metric     = "ROC",
                 tuneGrid   = expand.grid(k = seq(3, 31, by = 2)))

fit_knn$bestTune
```

The extra argument is `preProcess = c("center", "scale")`. kNN measures
distances, so features on larger numeric scales would dominate. It matters
that caret does this **inside each fold** — scaling on the whole dataset
first would leak information from the validation fold into training (the
preprocessing-leakage pitfall).

**12.**

```{r ex12}
fit_rf <- train(Class ~ ., data = train,
                method    = "rf",
                trControl = ctrl,
                metric    = "ROC",
                tuneGrid  = expand.grid(mtry = 2:6))

fit_rf$bestTune
```

**13.**

```{r ex13}
res <- resamples(list(tree = fit_tree_cv, knn = fit_knn, rf = fit_rf))
summary(res)
bwplot(res, metric = "ROC")
```

`resamples()` compares models on the **same folds**, so differences are
paired and fair. Read the mean ROC row for the winner — but also look at the
spread: if the boxplots overlap heavily, the "winner" may be within noise,
and simpler/more interpretable models become a reasonable choice. *(On this
dataset all three score very high AUC because BreastCancer is easy to
separate — a good moment to note that a high number is not automatically
impressive.)*

---

# Part 6 · Opening the test set

**14.**

```{r ex14}
# pick the winner by mean CV AUC, programmatically
cv_auc <- sapply(list(tree = fit_tree_cv, knn = fit_knn, rf = fit_rf),
                 function(m) max(m$results$ROC))
best_name  <- names(which.max(cv_auc))
best_model <- list(tree = fit_tree_cv, knn = fit_knn, rf = fit_rf)[[best_name]]
best_name

# ---- test set opened ONCE ----
final_pred <- predict(best_model, newdata = test)
confusionMatrix(final_pred, test$Class, positive = "malignant")

final_prob <- predict(best_model, newdata = test, type = "prob")[, "malignant"]
final_roc  <- roc(test$Class, final_prob, levels = c("benign", "malignant"))
auc(final_roc)                       # test AUC
max(best_model$results$ROC)          # CV AUC, for comparison
```

CV AUC and test AUC should be close. A large gap (test much lower than CV)
signals overfitting to the training data, or leakage that inflated the CV
estimate — the model won't reproduce in a new hospital. The gap almost always
goes one way: the estimate you tuned on is optimistic, the one you didn't touch
is honest.

**15.**

a. `Id` is a per-patient identifier — a form of **data leakage / a leaky
feature**. A model that keys off IDs memorizes *these specific patients*
rather than learning tumor biology; it looks excellent on data containing
those IDs and collapses on any new cohort, whose IDs it has never seen.

b. The colleague broke the golden rule: **never tune (or model-select) on
the test set.** By choosing the model with the best test AUC, the test set
became part of model selection — so the published AUC is optimistically
biased and will not hold on genuinely new patients. The honest approach:
select with cross-validation on the training data, then touch the test set
once.
