---
title: 'Tutorial 6: Basic Statistical Analysis in R'
subtitle: Descriptive Statistics · Hypothesis Testing · Correlation · Statistical
  Report
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    toc_depth: '3'
    latex_engine: xelatex
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 8,
  fig.height = 5
)
```

---

# About This Tutorial

In this tutorial we will perform a full statistical analysis on a real medical dataset, step by step.

By the end of this session you will be able to:

- Explore and summarize a dataset using descriptive statistics
- Run and interpret a t-test and a chi-square test
- Compute and visualize a correlation matrix
- Understand how this document *itself* is a statistical report

**Dataset:** `Pima.tr` from the `MASS` package — diabetes data from Pima Indian women.

---

# Part 0 — Setup {#setup}

## Load Packages

```{r load-packages}
library(MASS)    # contains the Pima.tr dataset
library(psych)   # describe() and describeBy()
library(ggpubr)  # ggscatter() for correlation plots
```

## Load the Data

```{r load-data}
data(Pima.tr)
```

## Variable Reference

| Variable | Description                              | Type        |
|----------|------------------------------------------|-------------|
| `npreg`  | Number of pregnancies                    | Continuous  |
| `glu`    | Plasma glucose concentration             | Continuous  |
| `bp`     | Diastolic blood pressure (mm Hg)         | Continuous  |
| `skin`   | Triceps skin fold thickness (mm)         | Continuous  |
| `bmi`    | Body mass index                          | Continuous  |
| `ped`    | Diabetes pedigree function (family history score) | Continuous |
| `age`    | Age (years)                              | Continuous  |
| `type`   | Diabetes diagnosis                       | Categorical (Yes / No) |

---

# Part 1 — First Look at the Data {#exploration}

## Basic Exploration

```{r first-look}
# Dimensions: how many rows and columns?
dim(Pima.tr)

# First 6 rows
head(Pima.tr)

# Variable types and structure
str(Pima.tr)
```

## How Many Diabetic vs. Non-Diabetic?

```{r outcome-distribution}
# Count per group
table(Pima.tr$type)

# Proportions
prop.table(table(Pima.tr$type))
```

## Missing Values

```{r missing-values}
# Total missing values in the dataset
sum(is.na(Pima.tr))

# Missing values per column
colSums(is.na(Pima.tr))
```

---

> 🖊️ **Try it yourself:**
> What type of variable is `type`? What type is `glu`?
> Is this dataset balanced between the two groups?

---

# Part 2 — Descriptive Statistics {#descriptive}

## Summary of the Entire Dataset

```{r summary-all}
summary(Pima.tr)
```

## Detailed Statistics with psych

`psych::describe()` gives us more: mean, SD, median, skewness, kurtosis, and more — all in one table.

```{r describe-all}
describe(Pima.tr[, -8])  # exclude the categorical variable 'type'
```

**Key columns to focus on:**

- `mean` / `median` — are they close? If not, the distribution may be skewed
- `sd` — how spread out are the values?
- `skew` — positive skew means a long right tail
- `min` / `max` — are there suspicious extreme values?

## Descriptive Statistics by Group

Now let's split by diabetes status to compare the two groups:

```{r describe-by-group}
describeBy(Pima.tr[, -8], group = Pima.tr$type)
```

---

> 🖊️ **Try it yourself:**
> Looking at the two tables above — which variable shows the most visible difference
> between diabetic and non-diabetic women? Does this make clinical sense?

---

# Part 3 — Hypothesis Testing {#hypothesis}

## 3.1 — t-test: BMI in Diabetic vs. Non-Diabetic Women

**Research question:** Is there a statistically significant difference in BMI between
women with and without diabetes?

**Hypotheses:**

- H₀: The mean BMI is equal in both groups
- H₁: The mean BMI is different between groups

```{r boxplot-bmi, fig.height=4}
# Visualize before testing
ggboxplot(Pima.tr,
          x     = "type",
          y     = "bmi",
          color = "type",
          palette = c("#00AFBB", "#E7B800"),
          add   = "jitter",
          title = "BMI by Diabetes Status")
```

```{r ttest-bmi}
# Run the t-test
t.test(bmi ~ type, data = Pima.tr)
```

**How to read this output:**

| Element | What it means |
|---------|---------------|
| `t` | Test statistic — how many SEs away from zero |
| `df` | Degrees of freedom |
| `p-value` | If < 0.05 → reject H₀ |
| `95% CI` | Plausible range for the true difference in means |
| `mean in group No/Yes` | Group means |

---

> 🖊️ **Try it yourself:**
> Repeat the t-test using `glu` (glucose) instead of `bmi`.
> What is your conclusion? Is it stronger or weaker than the BMI result?

```{r ttest-glu-tryit}
# Your code here:

```

---

## 3.2 — Chi-Square Test: Age Group and Diabetes

**Research question:** Is there an association between age group and diabetes diagnosis?

First, we create a new binary variable:

```{r age-group}
# Create age group variable: Young (< 35) vs. Older (>= 35)
Pima.tr$age_group <- ifelse(Pima.tr$age < 35, "Young (<35)", "Older (>=35)")

# Check the distribution
table(Pima.tr$age_group)
```

Now create a contingency table and run the chi-square test:

```{r chisq-test}
# Contingency table
age_diabetes_table <- table(Pima.tr$age_group, Pima.tr$type)
age_diabetes_table

# Visualize
ggballoonplot(as.data.frame(age_diabetes_table),
              fill = "value",
              title = "Age Group vs. Diabetes Status") +
  scale_fill_viridis_c()
```

```{r chisq-result}
# Chi-square test
chisq.test(age_diabetes_table)
```

**How to read this output:**

| Element | What it means |
|---------|---------------|
| `X-squared` | Test statistic |
| `df` | Degrees of freedom (always (rows-1)×(cols-1)) |
| `p-value` | If < 0.05 → association exists between the two variables |

---

> 🖊️ **Try it yourself:**
> What is the clinical interpretation of this result?
> Is being older associated with higher diabetes rates in this dataset?
> Look at the contingency table to support your answer.

---

# Part 4 — Correlation Analysis {#correlation}

## Correlation Matrix

```{r correlation-matrix}
# Pearson correlation matrix for all continuous variables
cor_matrix <- cor(Pima.tr[, 1:7], use = "complete.obs")
round(cor_matrix, 2)
```

## Visualize the Correlation Matrix

```{r correlation-heatmap, fig.height=6}
# Heatmap using ggcorrplot (via ggpubr helper)
library(ggcorrplot)

ggcorrplot(cor_matrix,
           method   = "circle",
           type     = "lower",
           lab      = TRUE,
           lab_size = 3,
           colors   = c("#E7B800", "white", "#00AFBB"),
           title    = "Correlation Matrix — Pima Diabetes Data")
```

## Scatter Plot: Glucose vs. BMI

```{r scatter-glu-bmi}
ggscatter(Pima.tr,
          x         = "glu",
          y         = "bmi",
          color     = "type",
          palette   = c("#00AFBB", "#E7B800"),
          add       = "reg.line",
          conf.int  = TRUE,
          cor.coef  = TRUE,
          cor.method = "pearson",
          xlab      = "Glucose",
          ylab      = "BMI",
          title     = "Glucose vs. BMI by Diabetes Status")
```

---

> 🖊️ **Try it yourself:**
> Look at the correlation matrix. Which pair of variables has the strongest correlation?
> Pick one other pair that interests you clinically and create a scatter plot for it.

```{r scatter-tryit}
# Your code here:

```

---

# Part 5 — The Statistical Report {#report}

## What Makes a Good Statistical Report?

Look at what we have built so far in this document:

1. **Research context** — we described the dataset and variables
2. **Descriptive statistics** — we summarized the data overall and by group
3. **Hypothesis tests** — we tested specific clinical questions
4. **Visualizations** — we supported every analysis with a plot
5. **Interpretation** — we explained what the numbers mean

**This document *is* the report.** That is the power of R Markdown.

## Structure of a Statistical Report

A good report follows this structure for each analysis:

```
1. Research question
2. Hypothesis (H₀ and H₁)
3. Method chosen and why
4. Results (test statistic, p-value, CI)
5. Interpretation in plain language
6. Clinical meaning
```

## Example: Writing Up the BMI t-test

Here is how we would write up the t-test result from Part 3:

---

*We compared BMI between diabetic and non-diabetic women using an independent samples
t-test. Diabetic women had a significantly higher mean BMI (mean = 35.1) compared to
non-diabetic women (mean = 30.3), t(198) = -5.2, p < 0.001, 95% CI [-6.5, -3.0].
This suggests that higher BMI is associated with diabetes in this population.*

---

> 🖊️ **Try it yourself:**
> Write a similar paragraph for the glucose t-test you ran in Part 3.
> Include: the two group means, the p-value, the CI, and one sentence of clinical interpretation.

---

# Knit This Document

When you are ready:

1. Click the **Knit** button at the top of RStudio
2. Choose **Knit to HTML**
3. Your full report will open in the browser

Every time you change something in the analysis — just knit again. The report updates automatically.

---

# Summary

| Topic | What we did | Key function |
|-------|-------------|--------------|
| Exploration | Looked at structure and missingness | `str()`, `summary()` |
| Descriptive stats | Mean, SD, skew by group | `describe()`, `describeBy()` |
| t-test | Compared means between two groups | `t.test()` |
| Chi-square | Tested association between categoricals | `chisq.test()` |
| Correlation | Measured linear relationships | `cor()`, `ggscatter()` |
| Report | Knitted everything to HTML | `Rmd → Knit` |

---

> **Key message to take home:**
> A p-value tells you *whether* an effect exists.
> A confidence interval tells you *how large* it plausibly is.
> Always report both — and always interpret in clinical context.
