---
title: "Week 4 Homework SOLUTIONS: Descriptive Statistics & Hypothesis Testing"
author: "Einat Borohovich"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Instructions

This is the **SOLUTION FILE** for Week 4 homework.

**Topics covered:**
- Descriptive statistics: mean, median, mode, variance, standard deviation
- Hypothesis testing: t-tests (one-sample, two-sample, paired) and chi-square tests
- Checking t-test assumptions

---

# Dataset: Emergency Room Patient Data

We'll use a dataset of patients admitted to the emergency room.

```{r}
# Load required libraries
library(ggplot2)
library(dplyr)

# Load the data
set.seed(123)
er_data <- data.frame(
  patient_id = 1:80,
  age = round(rnorm(80, mean = 45, sd = 15)),
  heart_rate = round(rnorm(80, mean = 85, sd = 15)),
  temperature = round(rnorm(80, mean = 37.5, sd = 1.2), 1),
  gender = sample(c("Male", "Female"), 80, replace = TRUE),
  smoker = sample(c("Yes", "No"), 80, replace = TRUE, prob = c(0.25, 0.75)),
  department = sample(c("Cardiology", "General"), 80, replace = TRUE)
)

# View first few rows
head(er_data)
```

---

# Part 1: Descriptive Statistics

## Exercise 1.1: Calculate Mean

Calculate the mean heart rate for all patients.

```{r}
# Calculate mean heart rate
mean_hr <- mean(er_data$heart_rate)
mean_hr
```

**Question:** What is the average heart rate? **`r round(mean_hr, 2)` bpm**

---

## Exercise 1.2: Calculate Median

Calculate the median temperature for all patients.

```{r}
# Calculate median temperature
median_temp <- median(er_data$temperature)
median_temp
```

**Question:** What is the median temperature? **`r median_temp`°C**

---

## Exercise 1.3: Find the Mode

Find the most common gender in the dataset.

```{r}
# Create a frequency table for gender
gender_table <- table(er_data$gender)
gender_table

# Identify which gender appears most often
mode_gender <- names(which.max(gender_table))
mode_gender
```

**Question:** What is the most common gender? **`r mode_gender`** (appears `r max(gender_table)` times)

---

## Exercise 1.4: Calculate Variance and Standard Deviation 

Calculate the variance and standard deviation of age.

```{r}
# Calculate variance of age
var_age <- var(er_data$age)
var_age

# Calculate standard deviation of age
sd_age <- sd(er_data$age)
sd_age

# Verify that SD = sqrt(variance)
sqrt(var_age)
sd_age
```

**Questions:**
- Variance = **`r round(var_age, 2)`**
- Standard Deviation = **`r round(sd_age, 2)`**
- Are they related as expected (SD = √variance)? **Yes, `r round(sd_age, 2)` = √`r round(var_age, 2)`**

---

## Exercise 1.5: Visualize with Histogram

Create a histogram of heart rate showing the mean and median as vertical lines.

```{r}
# Calculate mean and median heart rate first
mean_hr <- mean(er_data$heart_rate)
median_hr <- median(er_data$heart_rate)

# Create histogram with ggplot2
ggplot(er_data, aes(x = heart_rate)) +
  geom_histogram(bins = 15, fill = "steelblue", alpha = 0.7, color = "white") +
  geom_vline(xintercept = mean_hr, color = "red", 
             linetype = "dashed", linewidth = 1.2) +
  geom_vline(xintercept = median_hr, color = "darkgreen", 
             linetype = "solid", linewidth = 1.2) +
  annotate("text", x = mean_hr + 8, y = 12, 
           label = paste("Mean =", round(mean_hr, 1)), 
           color = "red", size = 4, fontface = "bold") +
  annotate("text", x = median_hr - 8, y = 10, 
           label = paste("Median =", round(median_hr, 1)), 
           color = "darkgreen", size = 4, fontface = "bold") +
  labs(title = "Distribution of Heart Rate",
       x = "Heart Rate (bpm)", 
       y = "Number of Patients") +
  theme_minimal()
```

**Interpretation:** The histogram shows the distribution of heart rates. Mean = `r round(mean_hr, 1)` bpm, Median = `r round(median_hr, 1)` bpm. They are very close, suggesting a relatively symmetric distribution.

---

## Exercise 1.6: Create a Boxplot

Create a boxplot comparing heart rate by gender.

```{r}
# Create boxplot
ggplot(er_data, aes(x = gender, y = heart_rate, fill = gender)) +
  geom_boxplot(alpha = 0.7) +
  labs(title = "Heart Rate by Gender",
       x = "Gender",
       y = "Heart Rate (bpm)") +
  theme_minimal() +
  theme(legend.position = "none")
```

**Interpretation:** The boxplots show the distribution of heart rate for each gender, including median, quartiles, and any outliers.

---

# Part 2: Checking T-Test Assumptions

Before running t-tests on heart rate data, check the assumptions.

## Exercise 2.1: Check if Data is Continuous 

```{r}
# Check the data type of heart_rate
class(er_data$heart_rate)

# Look at summary statistics
summary(er_data$heart_rate)
```

**Question:** Is heart_rate continuous? **Yes, it is numeric data (integer values representing continuous measurements).**

---

## Exercise 2.2: Check Normality

Create a histogram and Q-Q plot to check if heart rate is approximately normal.

```{r}
# Create histogram of heart_rate
ggplot(er_data, aes(x = heart_rate)) +
  geom_histogram(bins = 15, fill = "steelblue", alpha = 0.7, color = "white") +
  labs(title = "Check for Normality: Histogram",
       subtitle = "Should be roughly bell-shaped",
       x = "Heart Rate (bpm)", 
       y = "Count") +
  theme_minimal()

# Create Q-Q plot
ggplot(er_data, aes(sample = heart_rate)) +
  stat_qq(color = "steelblue", size = 2) +
  stat_qq_line(color = "red", linetype = "dashed", linewidth = 1) +
  labs(title = "Check for Normality: Q-Q Plot",
       subtitle = "Points should fall along the red line",
       x = "Theoretical Quantiles", 
       y = "Sample Quantiles") +
  theme_minimal()

# Shapiro-Wilk test
shapiro.test(er_data$heart_rate)
```

**Question:** Does the data appear approximately normal? **Yes, the histogram is roughly bell-shaped and the Q-Q plot points fall reasonably along the line. The Shapiro-Wilk test p-value is `r round(shapiro.test(er_data$heart_rate)$p.value, 3)`, which is `r ifelse(shapiro.test(er_data$heart_rate)$p.value > 0.05, "> 0.05 (normal)", "< 0.05 (not perfectly normal, but with n=80, Central Limit Theorem applies)")`. Additionally, n = 80 > 30, so t-test is appropriate.**

---

## Exercise 2.3: Check Independence 

```{r}
# Check if all patient_ids are unique
length(unique(er_data$patient_id))
nrow(er_data)

# Are they equal?
length(unique(er_data$patient_id)) == nrow(er_data)
```

**Question:** Are observations independent? **Yes, each patient has a unique ID (80 unique IDs for 80 rows), so observations are independent.**

---

# Part 3: Hypothesis Testing - T-Tests 

## Exercise 3.1: One-Sample T-Test 

Test if the average temperature is **higher than** normal body temperature (37.0°C).

```{r}
# Perform one-sample t-test
# H0: mean temperature ≤ 37.0
# H1: mean temperature > 37.0
t_test_temp <- t.test(er_data$temperature, 
                      mu = 37.0, 
                      alternative = "greater")

t_test_temp
```

**Questions:**
- p-value = **`r format.pval(t_test_temp$p.value, digits = 3)`**
- Is temperature significantly higher than 37.0°C? (p < 0.05) **`r ifelse(t_test_temp$p.value < 0.05, "Yes", "No")`, since p `r ifelse(t_test_temp$p.value < 0.05, "< 0.05", "≥ 0.05")` we `r ifelse(t_test_temp$p.value < 0.05, "reject", "fail to reject")` the null hypothesis.**

**Interpretation:** The mean temperature in our sample is `r round(t_test_temp$estimate, 2)`°C. `r ifelse(t_test_temp$p.value < 0.05, "This is significantly higher than normal body temperature (37.0°C).", "This is not significantly higher than normal body temperature (37.0°C).")`

---

## Exercise 3.2: Two-Sample T-Test 

Test if heart rate differs between males and females.

```{r}
# Before testing, check if variances are similar
# Calculate variance by gender
er_data %>%
  group_by(gender) %>%
  summarize(
    n = n(),
    mean_hr = mean(heart_rate),
    var_hr = var(heart_rate),
    sd_hr = sd(heart_rate)
  )

# Perform two-sample t-test
# H0: mean_male = mean_female
# H1: mean_male ≠ mean_female
t_test_gender <- t.test(heart_rate ~ gender, data = er_data)

t_test_gender
```

**Questions:**
- p-value = **`r format.pval(t_test_gender$p.value, digits = 3)`**
- Do males and females have significantly different heart rates? **`r ifelse(t_test_gender$p.value < 0.05, "Yes", "No")`, since p `r ifelse(t_test_gender$p.value < 0.05, "< 0.05", "≥ 0.05")` we `r ifelse(t_test_gender$p.value < 0.05, "reject", "fail to reject")` the null hypothesis.**

**Interpretation:** `r ifelse(t_test_gender$p.value < 0.05, "There is a significant difference in heart rate between males and females.", "There is no significant difference in heart rate between males and females.")`

---

## Exercise 3.3: Paired T-Test

Create before/after data and test if heart rate changed after treatment.

```{r}
# Create paired data (first 40 patients, measured twice)
set.seed(456)
treatment_data <- data.frame(
  patient_id = 1:40,
  hr_before = round(rnorm(40, mean = 95, sd = 12)),
  hr_after = round(rnorm(40, mean = 85, sd = 12))
)

# Perform paired t-test
# H0: mean difference = 0
# H1: mean difference ≠ 0
t_test_paired <- t.test(treatment_data$hr_before, 
                        treatment_data$hr_after, 
                        paired = TRUE)

t_test_paired

# Calculate mean difference
mean_diff <- mean(treatment_data$hr_before - treatment_data$hr_after)
cat("\nMean difference (before - after):", round(mean_diff, 2), "bpm")
```

**Questions:**
- p-value = **`r format.pval(t_test_paired$p.value, digits = 3)`**
- Did heart rate significantly change after treatment? **`r ifelse(t_test_paired$p.value < 0.05, "Yes", "No")`, since p `r ifelse(t_test_paired$p.value < 0.05, "< 0.05", "≥ 0.05")` we `r ifelse(t_test_paired$p.value < 0.05, "reject", "fail to reject")` the null hypothesis.**

**Interpretation:** `r ifelse(t_test_paired$p.value < 0.05, paste("Heart rate significantly decreased after treatment by an average of", round(mean_diff, 2), "bpm."), "Heart rate did not significantly change after treatment.")`

---

# Part 4: Chi-Square Test

## Exercise 4.1: Chi-Square Test

Test if smoking status is associated with department.

```{r}
# Create contingency table
cont_table <- table(er_data$smoker, er_data$department)
cont_table

# Add row and column totals for clarity
addmargins(cont_table)

# Perform chi-square test
chi_test <- chisq.test(cont_table)
chi_test

# Check expected counts (should all be ≥ 5)
chi_test$expected
```

**Questions:**
- p-value = **`r format.pval(chi_test$p.value, digits = 3)`**
- Is there a significant association between smoking and department? **`r ifelse(chi_test$p.value < 0.05, "Yes", "No")`, since p `r ifelse(chi_test$p.value < 0.05, "< 0.05", "≥ 0.05")` we `r ifelse(chi_test$p.value < 0.05, "reject", "fail to reject")` the null hypothesis.**

**Interpretation:** `r ifelse(chi_test$p.value < 0.05, "There is a significant association between smoking status and department - the distribution of smokers differs across departments.", "There is no significant association between smoking status and department - smoking prevalence is similar across departments.")`

**Note:** All expected counts are ≥ 5, so the chi-square test is appropriate.

---

# Reflection Questions (Bonus)

1. **When would you use median instead of mean?**

**Answer:** Use median when:
- Data is skewed (not symmetric)
- Data has outliers that would pull the mean away from the center
- Example: Income data, house prices, or any data with extreme values

The median is more robust to outliers because it only depends on the middle value, not all values.

2. **What should you do if t-test assumptions are violated?**

**Answer:** If t-test assumptions are violated:
- **Transform the data** (e.g., log transformation) to make it more normal
- **Use non-parametric tests** instead:
  - Instead of one-sample t-test → Wilcoxon signed-rank test
  - Instead of two-sample t-test → Mann-Whitney U test
  - Instead of paired t-test → Wilcoxon signed-rank test for paired data
- **Increase sample size** if possible (Central Limit Theorem helps with normality when n > 30)

3. **What's the difference between a one-sample and two-sample t-test?**

**Answer:**
- **One-sample t-test:** Compares the mean of ONE sample to a known value (e.g., is average BP different from 120 mmHg?)
- **Two-sample t-test:** Compares the means of TWO independent groups (e.g., do males and females have different BP?)
- **Paired t-test:** Special case - compares TWO measurements from the SAME subjects (e.g., BP before vs after treatment in the same patients)

---

# Summary of Results

```{r}
# Create summary table
results <- data.frame(
  Test = c("One-sample t-test", "Two-sample t-test", "Paired t-test", "Chi-square test"),
  Question = c(
    "Temp > 37.0°C?",
    "HR differs by gender?",
    "HR changed after treatment?",
    "Smoking associated with dept?"
  ),
  P_Value = c(
    round(t_test_temp$p.value, 4),
    round(t_test_gender$p.value, 4),
    round(t_test_paired$p.value, 4),
    round(chi_test$p.value, 4)
  ),
  Significant = c(
    ifelse(t_test_temp$p.value < 0.05, "Yes", "No"),
    ifelse(t_test_gender$p.value < 0.05, "Yes", "No"),
    ifelse(t_test_paired$p.value < 0.05, "Yes", "No"),
    ifelse(chi_test$p.value < 0.05, "Yes", "No")
  )
)

knitr::kable(results, caption = "Summary of Hypothesis Tests")
```



**Great work!** 🎉
