---
title: "Week 4: Descriptive Statistics & Hypothesis Testing"
author: "Einat Borohovich"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: show
    theme: readable
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Welcome!

---

# Part 1: Descriptive Statistics

## Setup: Creating Sample Data

Let's create a dataset of patient blood pressure readings:

```{r}
# Set a random seed so results are reproducible
set.seed(42)

# Create a data frame with simulated patient data
bp_data <- data.frame(
  # Create patient IDs from 1 to 100
  patient_id = 1:100,
  # Generate 100 ages from normal distribution (mean=55, sd=15), rounded to whole numbers
  age = round(rnorm(100, mean = 55, sd = 15)),
  # Generate 100 systolic BP values from normal distribution (mean=135, sd=20)
  systolic_bp = round(rnorm(100, mean = 135, sd = 20)),
  # Generate 100 diastolic BP values from normal distribution (mean=85, sd=12)
  diastolic_bp = round(rnorm(100, mean = 85, sd = 12)),
  # Randomly assign diabetes status: 30% "Yes", 70% "No"
  diabetes = sample(c("Yes", "No"), 100, replace = TRUE, prob = c(0.3, 0.7)),
  # Randomly assign department with equal probability
  department = sample(c("ER", "Cardiology", "Internal"), 100, replace = TRUE)
)

# Display the first 6 rows of the data frame
head(bp_data)
```

---

## Mean (Average)

The **mean** is the sum of all values divided by the count.

**Formula:** $\bar{x} = \frac{\sum x_i}{n}$

```{r}
# Calculate the mean (average) of systolic BP
mean_systolic <- mean(bp_data$systolic_bp)

# Print the result
mean_systolic

# If you had missing values (NA), you would use:
# mean(bp_data$systolic_bp, na.rm = TRUE)
# The na.rm = TRUE argument removes NAs before calculating
```

**Clinical interpretation:** The average systolic BP in our sample is `r round(mean_systolic, 1)` mmHg.

### Calculating Mean by Hand

Let's verify by calculating manually:

```{r}
# Sum all values
total <- sum(bp_data$systolic_bp)

# Count number of values
n <- length(bp_data$systolic_bp)

# Divide sum by count
manual_mean <- total / n

# Compare with mean() function
cat("mean() function:", mean_systolic, "\n")
cat("Manual calculation:", manual_mean, "\n")
cat("Same result?", mean_systolic == manual_mean)
```

---

## Median (Middle Value)

The **median** is the middle value when data is sorted from smallest to largest.

**For odd n:** middle value  
**For even n:** average of two middle values

```{r}
# Calculate the median (middle value) of systolic BP
median_systolic <- median(bp_data$systolic_bp)

# Print the result
median_systolic
```

### Understanding Median vs Mean

```{r}
# Compare mean vs median
cat("Mean:", round(mean_systolic, 2), "\n")
cat("Median:", round(median_systolic, 2), "\n")
cat("Difference:", round(mean_systolic - median_systolic, 2), "\n")
```

**When to use which:**

- **Mean:** Use when data is symmetric and has no outliers
- **Median:** Use when data is skewed or has outliers

### Example: Effect of Outliers

```{r}
# Create a small dataset to show outlier effect
small_data <- c(120, 125, 130, 135, 140)  # Symmetric data

# Add an outlier
data_with_outlier <- c(120, 125, 130, 135, 200)  # One very high value

# Compare mean and median
cat("Without outlier:\n")
cat("  Mean:", mean(small_data), "\n")
cat("  Median:", median(small_data), "\n\n")

cat("With outlier (200):\n")
cat("  Mean:", mean(data_with_outlier), "\n")
cat("  Median:", median(data_with_outlier), "\n")

# The mean jumped from 130 to 142, but median stayed at 130!
```

---

## Mode (Most Common Value)

The **mode** is the value that appears most frequently.

R doesn't have a built-in mode function, so we calculate it using `table()`:

```{r}
# Create a frequency table of ages (counts how many times each age appears)
age_table <- table(bp_data$age)

# Print the frequency table
head(age_table, 10)  # Show first 10 ages

# Find which age appears most frequently
# which.max() returns the position of the maximum value
# names() extracts the age value at that position
# as.numeric() converts it from text to number
mode_age <- as.numeric(names(age_table)[which.max(age_table)])

# Print the mode
cat("Most common age:", mode_age, "years old\n")

# How many patients have this age?
cat("Number of patients with age", mode_age, ":", max(age_table), "\n")
```

### Mode for Categorical Variables

Mode is especially useful for categorical data:

```{r}
# Frequency table for diabetes status
table(bp_data$diabetes)

# Frequency table for department
table(bp_data$department)

# The mode for department is the one with highest count
dept_table <- table(bp_data$department)
mode_dept <- names(dept_table)[which.max(dept_table)]
cat("\nMost common department:", mode_dept)
```

---

## Variance

**Variance** measures how spread out the data is from the mean.

**Formula:** $s^2 = \frac{\sum (x_i - \bar{x})^2}{n-1}$

```{r}
# Calculate variance using the var() function
variance_systolic <- var(bp_data$systolic_bp)

# Print the variance
variance_systolic
```

**Problem with variance:** The units are squared (mmHg²), which is hard to interpret!

### Calculating Variance Manually

Let's see how variance is calculated step by step:

```{r}
# Step 1: Calculate the mean
mean_val <- mean(bp_data$systolic_bp)

# Step 2: Calculate deviation from mean for each value
deviations <- bp_data$systolic_bp - mean_val

# Step 3: Square each deviation
squared_deviations <- deviations^2

# Step 4: Sum all squared deviations
sum_squared_dev <- sum(squared_deviations)

# Step 5: Divide by (n-1)
n <- length(bp_data$systolic_bp)
manual_variance <- sum_squared_dev / (n - 1)

# Compare with var() function
cat("var() function:", variance_systolic, "\n")
cat("Manual calculation:", manual_variance, "\n")
cat("Same result?", round(variance_systolic, 10) == round(manual_variance, 10))
```

**Note:** We divide by (n-1) instead of n. This is called "Bessel's correction" and gives us an unbiased estimate of population variance.

---

## Standard Deviation

**Standard deviation (SD)** is the square root of variance. It has the same units as the original data.

**Formula:** $s = \sqrt{s^2}$

```{r}
# Calculate standard deviation using the sd() function
sd_systolic <- sd(bp_data$systolic_bp)

# Print the SD
sd_systolic

# Verify: SD is the square root of variance
cat("\nsd() function:", sd_systolic, "\n")
cat("sqrt(variance):", sqrt(variance_systolic), "\n")
cat("Same result?", sd_systolic == sqrt(variance_systolic))
```

**Clinical interpretation:** Systolic BP is `r round(mean_systolic, 1)` ± `r round(sd_systolic, 1)` mmHg.

### What Does Standard Deviation Tell Us?

For normally distributed data:

- About **68%** of values are within 1 SD of the mean
- About **95%** of values are within 2 SD of the mean
- About **99.7%** of values are within 3 SD of the mean

```{r}
# Calculate the ranges
one_sd_lower <- mean_systolic - sd_systolic
one_sd_upper <- mean_systolic + sd_systolic

cat("Mean ± 1 SD range:", round(one_sd_lower, 1), "to", round(one_sd_upper, 1), "mmHg\n")

# Count how many values fall within this range
within_one_sd <- sum(bp_data$systolic_bp >= one_sd_lower & 
                     bp_data$systolic_bp <= one_sd_upper)

# Calculate percentage
percentage <- (within_one_sd / length(bp_data$systolic_bp)) * 100

cat("Percentage within 1 SD:", round(percentage, 1), "%\n")
```

---

## All Statistics Together: summary()

The `summary()` function gives you multiple statistics at once:

```{r}
# Summary statistics for systolic BP
summary(bp_data$systolic_bp)

# Summary for entire dataset
summary(bp_data)
```

The summary shows:
- **Min:** Minimum value
- **1st Qu:** 25th percentile (first quartile)
- **Median:** Middle value
- **Mean:** Average
- **3rd Qu:** 75th percentile (third quartile)
- **Max:** Maximum value

---

## Visualizing Descriptive Statistics

### Histogram with Mean and Median

A histogram shows the distribution of data and helps us see:

- The shape of the distribution (symmetric, skewed, etc.)
- Where most values cluster
- How mean and median compare

```{r}
# Load the ggplot2 library for plotting
library(ggplot2)
if (!requireNamespace("rlang", quietly = TRUE)) {
  install.packages("rlang")
}

# Create a histogram showing the distribution
ggplot(bp_data, aes(x = systolic_bp)) +  # Set x-axis to systolic_bp
  # geom_histogram creates bars showing frequency of values
  # bins = 20 means divide data into 20 bins
  geom_histogram(bins = 20, fill = "steelblue", alpha = 0.7, color = "white") +
  # Add a vertical dashed red line at the mean
  geom_vline(xintercept = mean_systolic, color = "red", 
             linetype = "dashed", size = 1.2) +
  # Add a vertical solid green line at the median
  geom_vline(xintercept = median_systolic, color = "darkgreen", 
             linetype = "solid", size = 1.2) +
  # Add text label for the mean
  annotate("text", x = mean_systolic + 15, y = 15, 
           label = paste("Mean =", round(mean_systolic, 1)), 
           color = "red", size = 4, fontface = "bold") +
  # Add text label for the median
  annotate("text", x = median_systolic - 15, y = 13, 
           label = paste("Median =", round(median_systolic, 1)), 
           color = "darkgreen", size = 4, fontface = "bold") +
  # Set axis labels and title
  labs(title = "Distribution of Systolic Blood Pressure",
       subtitle = paste("SD =", round(sd_systolic, 1), "mmHg"),
       x = "Systolic BP (mmHg)", 
       y = "Number of Patients") +
  # Use a clean theme
  theme_minimal()
```

**What this shows:** The red dashed line (mean) and green solid line (median) are close together, indicating the data is fairly symmetric without major outliers.

### Histogram Showing Variance/SD

Visualizing what standard deviation means:

```{r}
# Create histogram with shaded regions showing ±1 SD from mean
ggplot(bp_data, aes(x = systolic_bp)) +
  # Add shaded rectangle showing mean ± 1 SD range
  # This captures approximately 68% of the data
  annotate("rect", 
           xmin = mean_systolic - sd_systolic, 
           xmax = mean_systolic + sd_systolic,
           ymin = 0, ymax = Inf, 
           alpha = 0.2, fill = "yellow") +
  # Create the histogram
  geom_histogram(bins = 20, fill = "steelblue", alpha = 0.7, color = "white") +
  # Add vertical line at the mean
  geom_vline(xintercept = mean_systolic, color = "red", 
             linetype = "dashed", size = 1.2) +
  # Add labels showing the SD range
  annotate("text", x = mean_systolic, y = 18,
           label = paste("Mean ± 1 SD\n[", 
                        round(mean_systolic - sd_systolic, 1), "-",
                        round(mean_systolic + sd_systolic, 1), "]"),
           color = "red", size = 3.5, fontface = "bold") +
  # Set labels
  labs(title = "Standard Deviation: Spread Around the Mean",
       subtitle = "Yellow region = Mean ± 1 SD (contains ~68% of data)",
       x = "Systolic BP (mmHg)", 
       y = "Number of Patients") +
  theme_minimal()
```

**What this shows:** The yellow shaded area represents the range of Mean ± 1 SD. About 68% of patients fall within this range.

### Boxplot

A boxplot provides a different view of the data distribution:

```{r}
# Calculate quartile values first
q1_val <- quantile(bp_data$systolic_bp, 0.25)
q3_val <- quantile(bp_data$systolic_bp, 0.75)

# Create the boxplot
ggplot(bp_data, aes(x = "", y = systolic_bp)) +
  geom_boxplot(fill = "lightblue", alpha = 0.7, width = 0.5) +
  
  # Add horizontal lines for mean and median
  geom_hline(yintercept = mean_systolic, color = "red", 
             linetype = "dashed", linewidth = 1) +
  geom_hline(yintercept = median_systolic, color = "darkgreen", 
             linetype = "solid", linewidth = 1) +
  
  # Add labels for Q1 and Q3
  annotate("text", x = 1.35, y = q3_val, 
           label = paste("Q3 (75%) =", round(q3_val, 1)), 
           color = "black", size = 4, hjust = 0) +
  annotate("text", x = 1.35, y = q1_val, 
           label = paste("Q1 (25%) =", round(q1_val, 1)), 
           color = "black", size = 4, hjust = 0) +
  
  # Add labels for mean and median with vertical adjustment (vjust) to avoid smudging
  annotate("text", x = 1.35, y = mean_systolic, 
           label = paste("Mean =", round(mean_systolic, 1)), 
           color = "red", fontface = "bold", size = 4, hjust = 0, vjust = -0.7) +
  annotate("text", x = 1.35, y = median_systolic, 
           label = paste("Median =", round(median_systolic, 1)), 
           color = "darkgreen", fontface = "bold", size = 4, hjust = 0, vjust = 1.7) +
  
  # Set titles and labels
  labs(title = "Boxplot of Systolic Blood Pressure",
       subtitle = "Displaying Mean, Median, and Quartiles (Q1/Q3)",
       y = "Systolic BP (mmHg)",
       x = "") +
  
  # Theme and layout adjustments
  theme_minimal() +
  theme(axis.text.x = element_blank(),
        plot.margin = margin(5, 100, 5, 5)) + # Extra right margin for labels
  coord_cartesian(clip = "off") # Prevent labels from being cut off
```

**Reading the boxplot:**

- **Bottom of box:** 25th percentile (Q1)
- **Line in box:** Median (50th percentile)
- **Top of box:** 75th percentile (Q3)
- **Whiskers:** Extend to min/max values
- **Dots (if any):** Potential outliers

---

---

# Part 2: Hypothesis Testing (45 minutes)

## What is Hypothesis Testing?

**Goal:** Determine if observed differences are real or just due to chance.

**Basic process:**

1. **State hypotheses:**
   - **Null hypothesis (H₀):** No difference exists
   - **Alternative hypothesis (H₁):** A difference exists

2. **Choose significance level:** Usually α = 0.05 (5%)

3. **Calculate test statistic** from your data

4. **Calculate p-value:** Probability of seeing your result if H₀ is true

5. **Make decision:**
   - If **p < 0.05**: Reject H₀ (result is statistically significant)
   - If **p ≥ 0.05**: Fail to reject H₀ (result is not statistically significant)

---

## T-Test Assumptions

**IMPORTANT:** Check these assumptions BEFORE running t-tests!

### 1. Data is Continuous

T-tests work on continuous numerical data (like blood pressure, weight, height), not categorical data (like gender, yes/no responses).

```{r}
# Check data type
class(bp_data$systolic_bp)  # Should be "numeric" or "integer"

# Look at the data
summary(bp_data$systolic_bp)  # Should show numeric values
```

### 2. Data is Approximately Normal

The data should follow a normal (bell-shaped) distribution, OR your sample size should be > 30.

```{r}
# Visual check: Histogram
# A normal distribution looks roughly bell-shaped and symmetric
ggplot(bp_data, aes(x = systolic_bp)) +
  geom_histogram(bins = 20, fill = "steelblue", alpha = 0.7, color = "white") +
  labs(title = "Check for Normality: Histogram",
       subtitle = "Should be roughly bell-shaped",
       x = "Systolic BP (mmHg)", y = "Count") +
  theme_minimal()

# Visual check: Q-Q plot
# Points should fall roughly along the diagonal line
ggplot(bp_data, aes(sample = systolic_bp)) +
  # stat_qq creates quantile-quantile plot
  stat_qq(color = "steelblue", size = 2) +
  # stat_qq_line adds the reference line
  stat_qq_line(color = "red", linetype = "dashed", size = 1) +
  labs(title = "Check for Normality: Q-Q Plot",
       subtitle = "Points should fall along the red line",
       x = "Theoretical Quantiles", y = "Sample Quantiles") +
  theme_minimal()

# Statistical test: Shapiro-Wilk test
# H0: Data is normally distributed
# If p > 0.05, data is approximately normal
shapiro.test(bp_data$systolic_bp)

# Note: With large samples (n > 30), small deviations from normality are okay
# The Central Limit Theorem helps us here!
cat("Sample size:", nrow(bp_data), "\n")
cat("If n > 30, t-test is robust to violations of normality")
```

### 3. Independence of Observations

Each observation should be independent - one measurement shouldn't influence another.

**Good examples (independent):**
- Blood pressure from different patients
- Test scores from different students

**Bad examples (NOT independent):**
- Multiple measurements from the same patient (use paired t-test instead!)
- Students in the same class who studied together

```{r}
# Check: Look at how data was collected
# Are these different patients? YES - they have unique patient IDs
length(unique(bp_data$patient_id))  # Should equal number of rows for independence
nrow(bp_data)

# If these numbers are equal, observations are independent
```

### 4. For Two-Sample T-Test: Similar Variances

When comparing two groups, their variances should be roughly similar (homogeneity of variance).

```{r}

library(dplyr)
# Calculate variance for each group
bp_data %>%
  # Split by diabetes status
  group_by(diabetes) %>%
  # Calculate variance for each group
  summarize(
    variance = var(systolic_bp),
    sd = sd(systolic_bp)
  )

# Rule of thumb: Largest variance should be < 4 times the smallest variance

# Formal test: Levene's test (from car package)
# Install if needed: install.packages("car")
library(car)

# H0: Variances are equal
# If p > 0.05, variances are similar enough
leveneTest(systolic_bp ~ diabetes, data = bp_data)

# If variances are unequal, use: t.test(..., var.equal = FALSE)
```

### What If Assumptions Are Violated?

If assumptions are not met, consider:

1. **Transform your data** (e.g., log transformation)
2. **Use non-parametric tests:**
   - Instead of one-sample t-test → Wilcoxon signed-rank test
   - Instead of two-sample t-test → Mann-Whitney U test (Wilcoxon rank-sum)
   - Instead of paired t-test → Wilcoxon signed-rank test for paired data

```{r}
# Example: Non-parametric alternative to two-sample t-test
# wilcox.test() doesn't assume normality
wilcox.test(systolic_bp ~ diabetes, data = bp_data)
```

**Summary of when to use t-tests:**
- ✅ Continuous data
- ✅ Approximately normal distribution OR n > 30
- ✅ Independent observations
- ✅ (For two-sample) Similar variances

---

## T-Tests: Comparing Means

T-tests are used when comparing means (averages).

### One-Sample T-Test

**Question:** Is our sample mean **higher than** a known value?

**Example:** Is the average systolic BP in our sample **higher than** normal (120 mmHg)?

```{r}
# H0: population mean ≤ 120 mmHg (not higher)
# H1: population mean > 120 mmHg (is higher)

# t.test() performs the one-sample t-test
# mu = 120 specifies the value we're testing against
# alternative = "greater" makes this a one-sided test
t_test_result <- t.test(bp_data$systolic_bp, 
                        mu = 120, 
                        alternative = "greater")

# Print the complete test results
t_test_result
```

**Understanding the output:**

- **t statistic:** `r round(t_test_result$statistic, 2)` - measures how many standard errors the sample mean is from 120
- **df (degrees of freedom):** `r t_test_result$parameter` - sample size minus 1
- **p-value:** `r format.pval(t_test_result$p.value, digits = 4)` - probability of this result if H₀ is true
- **95% confidence interval:** Lower bound = `r round(t_test_result$conf.int[1], 2)`. We're 95% confident the true mean is above this value.
- **sample mean:** `r round(t_test_result$estimate, 2)` mmHg

**Conclusion:** Since p-value is `r ifelse(t_test_result$p.value < 0.05, "< 0.05", "≥ 0.05")`, we `r ifelse(t_test_result$p.value < 0.05, "reject", "fail to reject")` the null hypothesis. The average BP in our sample `r ifelse(t_test_result$p.value < 0.05, "is significantly higher than", "is not significantly higher than")` 120 mmHg.

**Note on one-sided vs two-sided tests:**
- **One-sided (what we used):** Tests if mean is specifically HIGHER (or LOWER)
  - Use `alternative = "greater"` to test if mean > reference value
  - Use `alternative = "less"` to test if mean < reference value
- **Two-sided (default):** Tests if mean is DIFFERENT in either direction
  - Use `alternative = "two.sided"` or omit the argument

```{r}
# Example: Two-sided test (testing if different, not specifically higher)
t.test(bp_data$systolic_bp, mu = 120)  # Tests if mean ≠ 120
```

**Conclusion:** Since p-value `r ifelse(t_test_result$p.value < 0.05, "< 0.05", "≥ 0.05")`, we `r ifelse(t_test_result$p.value < 0.05, "reject", "fail to reject")` the null hypothesis. The average BP in our sample `r ifelse(t_test_result$p.value < 0.05, "is significantly different from", "is not significantly different from")` 120 mmHg.

```{r}
# Extract specific components
# Extract just the p-value
t_test_result$p.value

# Extract the 95% confidence interval
t_test_result$conf.int

# Extract the sample mean
t_test_result$estimate
```

---

### Two-Sample T-Test (Independent Groups)

**Question:** Do two independent groups have different means?

**Example:** Do diabetic patients have different systolic BP than non-diabetic patients?

```{r}
# H0: mean_diabetic = mean_non_diabetic
# H1: mean_diabetic ≠ mean_non_diabetic

# The formula syntax "outcome ~ group" compares outcome across groups
# data = bp_data tells R which dataset to use
t_test_diabetes <- t.test(systolic_bp ~ diabetes, data = bp_data)

# Print the results
t_test_diabetes
```

**Understanding the output:**

- **t statistic:** `r round(t_test_diabetes$statistic, 2)`
- **df:** `r round(t_test_diabetes$parameter, 2)`
- **p-value:** `r format.pval(t_test_diabetes$p.value, digits = 4)`
- **95% CI for difference:** [`r round(t_test_diabetes$conf.int[1], 2)`, `r round(t_test_diabetes$conf.int[2], 2)`]
- **Group means:** 
  - No: `r round(t_test_diabetes$estimate[1], 2)` mmHg
  - Yes: `r round(t_test_diabetes$estimate[2], 2)` mmHg

**Conclusion:** `r ifelse(t_test_diabetes$p.value < 0.05, "There is a statistically significant difference", "There is no statistically significant difference")` in systolic BP between diabetic and non-diabetic patients (p = `r round(t_test_diabetes$p.value, 4)`).

### Calculate Group Statistics Manually

```{r}
# Load dplyr for data manipulation
library(dplyr)

# Calculate summary statistics by diabetes group
group_summary <- bp_data %>%
  # Split data by diabetes status
  group_by(diabetes) %>%
  # Calculate mean, SD, and count for each group
  summarize(
    n = n(),                        # Count number in each group
    mean_bp = mean(systolic_bp),    # Mean BP for each group
    sd_bp = sd(systolic_bp)         # SD for each group
  )

# Print the summary table
print(group_summary)
```

### Visualize the Comparison

```{r}
# Create a boxplot comparing the two groups
ggplot(bp_data, aes(x = diabetes, y = systolic_bp, fill = diabetes)) +
  # Create boxplots
  geom_boxplot(alpha = 0.7, width = 0.5) +
  # Add individual data points with some horizontal jitter to avoid overlap
  geom_jitter(width = 0.15, alpha = 0.4, size = 2) +
  # Add red diamonds showing the mean for each group
  stat_summary(fun = mean, geom = "point", shape = 23, size = 4, 
               fill = "red", color = "darkred") +
  # Add labels with p-value in subtitle
  labs(title = "Systolic BP: Diabetic vs Non-Diabetic Patients",
       subtitle = paste("Two-sample t-test: p =", 
                       round(t_test_diabetes$p.value, 4)),
       x = "Diabetes Status", 
       y = "Systolic BP (mmHg)") +
  # Use a clean theme
  theme_minimal() +
  # Change legend title and remove legend (since x-axis labels are clear)
  theme(legend.position = "none") +
  # Use custom colors
  scale_fill_manual(values = c("lightblue", "lightcoral"))
```

---

### Paired T-Test

**Question:** Did the same subjects change over time or between conditions?

**Key difference:** Each subject serves as their own control (measurements are paired).

**Example:** Did blood pressure change in the same patients before vs after treatment?

```{r}
# Create paired data (same patients measured twice)
set.seed(123)  # Set seed for reproducibility
treatment_data <- data.frame(
  # Patient IDs 1 to 50
  patient_id = 1:50,
  # BP before treatment: mean 145, sd 15
  bp_before = round(rnorm(50, mean = 145, sd = 15)),
  # BP after treatment: mean 138, sd 15 (on average, 7 mmHg lower)
  bp_after = round(rnorm(50, mean = 138, sd = 15))
)

# Look at first few patients
head(treatment_data)

# H0: mean difference = 0 (no change in BP)
# H1: mean difference ≠ 0 (BP changed)

# paired = TRUE tells R these are matched measurements from same patients
t_test_paired <- t.test(treatment_data$bp_before, 
                        treatment_data$bp_after, 
                        paired = TRUE)

# Print results
t_test_paired
```

**Understanding paired t-test output:**

- **t statistic:** `r round(t_test_paired$statistic, 2)`
- **df:** `r t_test_paired$parameter` (n - 1, where n = number of pairs)
- **p-value:** `r format.pval(t_test_paired$p.value, digits = 4)`
- **mean difference:** `r round(t_test_paired$estimate, 2)` mmHg (before minus after)
- **95% CI for difference:** [`r round(t_test_paired$conf.int[1], 2)`, `r round(t_test_paired$conf.int[2], 2)`]

**Conclusion:** `r ifelse(t_test_paired$p.value < 0.05, "Blood pressure significantly changed", "Blood pressure did not significantly change")` after treatment (p = `r round(t_test_paired$p.value, 4)`).

### Visualize Paired Data

```{r}
# Calculate change for each patient (after minus before)
# Negative values = BP decreased (improvement)
treatment_data$change <- treatment_data$bp_after - treatment_data$bp_before

# Create histogram of changes
ggplot(treatment_data, aes(x = change)) +
  # Histogram of BP changes
  geom_histogram(bins = 15, fill = "coral", alpha = 0.7, color = "white") +
  # Dashed red line at zero (no change)
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", size = 1.2) +
  # Solid blue line at mean change
  geom_vline(xintercept = mean(treatment_data$change), 
             linetype = "solid", color = "darkblue", size = 1.2) +
  # Add label for mean change
  annotate("text", x = mean(treatment_data$change) + 5, y = 10,
           label = paste("Mean change =", round(mean(treatment_data$change), 1), "mmHg"),
           color = "darkblue", fontface = "bold", size = 4) +
  # Add title with p-value
  labs(title = "Change in BP After Treatment",
       subtitle = paste("Paired t-test: p =", round(t_test_paired$p.value, 4)),
       x = "BP Change (mmHg)\nNegative = Improvement", 
       y = "Number of Patients") +
  theme_minimal()
```

---


---

## Chi-Square Test: Categorical Variables

**When to use:** Testing association between two categorical variables.

**Question:** Are two categorical variables independent, or is there a relationship?

**Example:** Is diabetes prevalence different across departments?

### Create a Contingency Table

```{r}
# Create a contingency table (cross-tabulation)
# Rows = diabetes status, Columns = department
diabetes_dept_table <- table(bp_data$diabetes, bp_data$department)

# Print the table showing counts in each combination
diabetes_dept_table

# Add row and column totals for clarity
addmargins(diabetes_dept_table)
```

### Calculate Proportions

```{r}
# Calculate proportions by column (margin = 2 means by column)
# This shows what percentage of each department has diabetes
prop.table(diabetes_dept_table, margin = 2)

# Multiply by 100 to get percentages
round(prop.table(diabetes_dept_table, margin = 2) * 100, 1)
```

### Perform Chi-Square Test

```{r}
# H0: Diabetes and department are independent (no association)
# H1: Diabetes and department are related (there is an association)

# chisq.test() performs chi-square test of independence
chi_test <- chisq.test(diabetes_dept_table)

# Print the test results
chi_test
```

**Understanding the output:**

- **χ² (chi-square) statistic:** `r round(chi_test$statistic, 3)` - measures how far observed counts are from expected counts if variables were independent
- **df (degrees of freedom):** `r chi_test$parameter` - calculated as (rows - 1) × (columns - 1) = (2-1) × (3-1) = 2
- **p-value:** `r format.pval(chi_test$p.value, digits = 4)` - probability of seeing this pattern if variables are truly independent

**Conclusion:** `r ifelse(chi_test$p.value < 0.05, "There is a statistically significant association", "There is no statistically significant association")` between diabetes status and department (p = `r round(chi_test$p.value, 4)`).

### Expected Counts

Chi-square test assumes that expected counts in each cell are ≥ 5:

```{r warning=TRUE}
# Look at expected counts under the null hypothesis
chi_test$expected

# All cells have counts ≥ 5, so chi-square test is valid

# Create a small contingency table that will trigger chi-square warning
small_table <- table(bp_data$diabetes[1:20], bp_data$department[1:20])

# Display the table
small_table

# Perform chi-square test (will show warning)
chi_test <- chisq.test(small_table)
chi_test

# Show expected counts (some will be < 5)
cat("\nExpected counts:\n")
chi_test$expected

# When expected counts < 5, use Fisher's exact test instead
cat("\n--- Using Fisher's Exact Test (appropriate for small samples) ---\n")
fisher_test <- fisher.test(small_table)
fisher_test
```

### Visualize the Association

```{r}
# Convert contingency table to data frame for ggplot
diabetes_dept_df <- as.data.frame(diabetes_dept_table)

# Give the columns meaningful names
names(diabetes_dept_df) <- c("Diabetes", "Department", "Count")

# Create a grouped bar chart
ggplot(diabetes_dept_df, aes(x = Department, y = Count, fill = Diabetes)) +
  # position = "dodge" puts bars side-by-side instead of stacked
  geom_bar(stat = "identity", position = "dodge", alpha = 0.8) +
  # Add counts as labels on top of bars
  geom_text(aes(label = Count), 
            position = position_dodge(width = 0.9),
            vjust = -0.5, size = 3.5) +
  # Add chi-square test result to subtitle
  labs(title = "Diabetes Prevalence by Department",
       subtitle = paste("Chi-square test: χ² =", round(chi_test$statistic, 2),
                       ", p =", round(chi_test$p.value, 4)),
       y = "Number of Patients") +
  # Use custom colors
  scale_fill_manual(values = c("lightblue", "lightcoral")) +
  theme_minimal() +
  # Move legend to top
  theme(legend.position = "top")



## ANOVA: Comparing More Than Two Groups
# Question: Does systolic BP differ across all three departments?

# 1. Run the ANOVA model
# H0: mean_Cardiology = mean_ER = mean_Internal
anova_model <- aov(systolic_bp ~ department, data = bp_data)

# 2. View the results
summary(anova_model)

# 3. Post-hoc analysis (Run this if p-value < 0.05)
# It tells us which specific departments are different
TukeyHSD(anova_model)

library(ggplot2)

ggplot(bp_data, aes(x = department, y = systolic_bp, fill = department)) +
  geom_boxplot(alpha = 0.7) +
  # Add individual points to show distribution
  geom_jitter(width = 0.1, alpha = 0.3) +
  # Add a red diamond for the mean of each group
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, fill = "red") +
  labs(title = "Systolic BP by Department",
       subtitle = "ANOVA p = 0.073 (Not Significant)",
       x = "Hospital Department",
       y = "Systolic BP (mmHg)") +
  theme_minimal() +
  theme(legend.position = "none")
```

---

## Summary: Choosing the Right Test

### T-Tests (for comparing means)

| Test Type | When to Use | R Code |
|-----------|-------------|--------|
| **One-sample** | Compare sample mean to known value | `t.test(x, mu = value)` |
| **Two-sample** | Compare means of two independent groups | `t.test(outcome ~ group, data)` |
| **Paired** | Compare means of same subjects at two times | `t.test(before, after, paired = TRUE)` |

### Chi-Square Test (for categorical variables)

| Test Type | When to Use | R Code |
|-----------|-------------|--------|
| **Chi-square** | Test association between two categorical variables | `chisq.test(table(var1, var2))` |

---

## Key Takeaways

**Descriptive Statistics:**

✓ **Mean:** Average, sensitive to outliers  
✓ **Median:** Middle value, robust to outliers  
✓ **Mode:** Most common value  
✓ **Variance:** Average squared deviation from mean (σ²)  
✓ **Standard Deviation:** Square root of variance (σ), same units as data  

**Hypothesis Testing:**

✓ **P-value < 0.05** means result is statistically significant  
✓ **One-sample t-test:** Compare mean to known value  
✓ **Two-sample t-test:** Compare means of two independent groups  
✓ **Paired t-test:** Compare same subjects at two time points  
✓ **Chi-square test:** Test association between categorical variables  

**Best Practices:**

✓ Always visualize your data before testing  
✓ Check assumptions (normality for t-tests, expected counts ≥5 for chi-square)  
✓ Report both the test statistic and p-value  
✓ Statistical significance ≠ clinical importance  

---


```

---


Remember: Statistics help us make decisions, but understanding what the numbers mean is more important than memorizing formulas!

Good luck with your analyses! 🎉
