---
title: "Week 4 Homework: Descriptive Statistics & Hypothesis Testing"
author: "Einat Borohovich"
output:
  html_document:
    toc: true
    toc_float: true
    theme: readable
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Instructions

Complete the following exercises using the skills learned in Week 4. 

**Topics covered:**
- Descriptive statistics: mean, median, mode, variance, standard deviation
- Hypothesis testing: t-tests (one-sample, two-sample, paired) and chi-square tests
- Checking t-test assumptions

**How to complete:**
1. Fill in the code where you see comments starting with `# YOUR CODE HERE:`
2. Run each code chunk to check your work
3. Knit the document to HTML when finished

---

# Dataset: Emergency Room Patient Data

We'll use a dataset of patients admitted to the emergency room.

```{r}
# Load required libraries
library(ggplot2)
library(dplyr)

# Load the data
set.seed(123)
er_data <- data.frame(
  patient_id = 1:80,
  age = round(rnorm(80, mean = 45, sd = 15)),
  heart_rate = round(rnorm(80, mean = 85, sd = 15)),
  temperature = round(rnorm(80, mean = 37.5, sd = 1.2), 1),
  gender = sample(c("Male", "Female"), 80, replace = TRUE),
  smoker = sample(c("Yes", "No"), 80, replace = TRUE, prob = c(0.25, 0.75)),
  department = sample(c("Cardiology", "General"), 80, replace = TRUE)
)

# View first few rows
head(er_data)
```

---

# Part 1: Descriptive Statistics 

## Exercise 1.1: Calculate Mean 

Calculate the mean heart rate for all patients.

```{r}
# YOUR CODE HERE: Calculate mean heart rate


```

**Question:** What is the average heart rate? _____________

---

## Exercise 1.2: Calculate Median 

Calculate the median temperature for all patients.

```{r}
# YOUR CODE HERE: Calculate median temperature


```

**Question:** What is the median temperature? _____________

---

## Exercise 1.3: Find the Mode

Find the most common gender in the dataset.

```{r}
# YOUR CODE HERE: Create a frequency table for gender


# YOUR CODE HERE: Identify which gender appears most often


```

**Question:** What is the most common gender? _____________

---

## Exercise 1.4: Calculate Variance and Standard Deviation

Calculate the variance and standard deviation of age.

```{r}
# YOUR CODE HERE: Calculate variance of age


# YOUR CODE HERE: Calculate standard deviation of age


# YOUR CODE HERE: Verify that SD = sqrt(variance)


```

**Questions:**
- Variance = _____________
- Standard Deviation = _____________
- Are they related as expected (SD = √variance)? _____________

---

## Exercise 1.5: Visualize with Histogram

Create a histogram of heart rate showing the mean and median as vertical lines.

```{r}
# YOUR CODE HERE: Calculate mean and median heart rate first



# YOUR CODE HERE: Create histogram with ggplot2
# Include:
# - geom_histogram() with bins = 15
# - geom_vline() for mean (red, dashed)
# - geom_vline() for median (green, solid)
# - Appropriate labels




```

---

## Exercise 1.6: Create a Boxplot

Create a boxplot comparing heart rate by gender.

```{r}
# YOUR CODE HERE: Create boxplot
# x-axis = gender, y-axis = heart_rate




```

---

# Part 2: Checking T-Test Assumptions

Before running t-tests on heart rate data, check the assumptions.

## Exercise 2.1: Check if Data is Continuous

```{r}
# YOUR CODE HERE: Check the data type of heart_rate


# YOUR CODE HERE: Look at summary statistics


```

**Question:** Is heart_rate continuous? _____________

---

## Exercise 2.2: Check Normality 

Create a histogram and Q-Q plot to check if heart rate is approximately normal.

```{r}
# YOUR CODE HERE: Create histogram of heart_rate




# YOUR CODE HERE: Create Q-Q plot




```

**Question:** Does the data appear approximately normal? _____________

---

## Exercise 2.3: Check Independence

```{r}
# YOUR CODE HERE: Check if all patient_ids are unique



```

**Question:** Are observations independent? _____________

---

# Part 3: Hypothesis Testing - T-Tests 

## Exercise 3.1: One-Sample T-Test 

Test if the average temperature is **higher than** normal body temperature (37.0°C).

```{r}
# YOUR CODE HERE: Perform one-sample t-test
# H0: mean temperature ≤ 37.0
# H1: mean temperature > 37.0
# Use alternative = "greater"




```

**Questions:**
- p-value = _____________
- Is temperature significantly higher than 37.0°C? (p < 0.05) _____________

---

## Exercise 3.2: Two-Sample T-Test 

Test if heart rate differs between males and females.

```{r}
# YOUR CODE HERE: Before testing, check if variances are similar
# Calculate variance by gender



# YOUR CODE HERE: Perform two-sample t-test
# H0: mean_male = mean_female
# H1: mean_male ≠ mean_female




```

**Questions:**
- p-value = _____________
- Do males and females have significantly different heart rates? _____________

---

## Exercise 3.3: Paired T-Test 

Create before/after data and test if heart rate changed after treatment.

```{r}
# Create paired data (first 40 patients, measured twice)
set.seed(456)
treatment_data <- data.frame(
  patient_id = 1:40,
  hr_before = round(rnorm(40, mean = 95, sd = 12)),
  hr_after = round(rnorm(40, mean = 85, sd = 12))
)

# YOUR CODE HERE: Perform paired t-test
# H0: mean difference = 0
# H1: mean difference ≠ 0
# Use paired = TRUE




```

**Questions:**
- p-value = _____________
- Did heart rate significantly change after treatment? _____________

---

# Part 4: Chi-Square Test

## Exercise 4.1: Chi-Square Test

Test if smoking status is associated with department.

```{r}
# YOUR CODE HERE: Create contingency table



# YOUR CODE HERE: Perform chi-square test




```

**Questions:**
- p-value = _____________
- Is there a significant association between smoking and department? _____________

---

# Reflection Questions (Bonus)

1. When would you use median instead of mean?

**Your answer:** 

2. What should you do if t-test assumptions are violated?

**Your answer:** 

3. What's the difference between a one-sample and two-sample t-test?

**Your answer:** 

---



**Good luck!** 🎉
