---
title: "Homework: Classification & Dimensionality Reduction"
subtitle: "Decision Trees · Naïve Bayes · PCA"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 9,
  fig.height = 5,
  fig.align  = "center"
)
```

## Setup

```{r packages}
library(mlbench)
library(rpart)
library(rpart.plot)
library(e1071)
library(caret)
library(pROC)
library(ggplot2)
library(dplyr)
library(corrplot)
library(factoextra)

theme_set(theme_minimal(base_size = 14))
```

We'll use the **BreastCancer** dataset from `mlbench` — a different dataset than the one used in class. It contains 699 observations of breast mass biopsies with 9 features and a binary outcome: benign vs. malignant.

```{r load-data}
data(BreastCancer)
df <- BreastCancer

# Remove the ID column and drop rows with missing values
df <- df %>% select(-Id) %>% na.omit()

# Convert all feature columns from factor to numeric
df <- df %>%
  mutate(across(-Class, ~as.numeric(as.character(.))))

str(df)
```

```{r split}
set.seed(123)
idx   <- createDataPartition(df$Class, p = 0.75, list = FALSE)
train <- df[idx, ]
test  <- df[-idx, ]
```

---

## Part 1 — Data Exploration

### Q1.1

What is the outcome distribution in the training set? What is the prevalence of malignant cases? Is this dataset balanced or imbalanced?

```{r q1-1}
# Your code here

```

### Q1.2

Plot the density distributions of at least 3 features, colored by outcome (benign vs. malignant). Which features seem to separate the two groups best?

```{r q1-2}
# Your code here

```

---

## Part 2 — Decision Tree

### Q2.1

Build a decision tree to predict `Class` from all features. Plot the tree. What is the first variable it splits on? Why do you think the algorithm chose that variable?

```{r q2-1}
# Your code here

```

### Q2.2

Look at the cp table (`printcp`). What is the optimal cp value? Build a pruned tree using that value and plot it.

```{r q2-2}
# Your code here

```

### Q2.3

Generate predictions on the test set. What is the accuracy? What is the sensitivity? What is the specificity?

```{r q2-3}
# Your code here

```

---

## Part 3 — Naïve Bayes

### Q3.1

Build a Naïve Bayes model on the training set. Look at the model output — pick one feature and explain in your own words what the mean and standard deviation per class tell you.

```{r q3-1}
# Your code here

```

### Q3.2

Generate predictions on the test set. Compare the confusion matrix to the decision tree from Q2.3. Which model has higher sensitivity? Which has higher specificity?

```{r q3-2}
# Your code here

```

### Q3.3

Create a new observation with the following values and predict its class with both models (decision tree and Naïve Bayes). Do they agree?

- Cl.thickness = 8, Cell.size = 5, Cell.shape = 5, Marg.adhesion = 3, Epith.c.size = 4, Bare.nuclei = 8, Bl.cromatin = 5, Normal.nucleoli = 5, Mitoses = 2

```{r q3-3}
# Your code here

```

---

## Part 4 — Model Evaluation

### Q4.1

Plot the ROC curves for both models (decision tree and Naïve Bayes) on the same plot. Report the AUC for each. Which model discriminates better overall?

```{r q4-1}
# Your code here

```

### Q4.2

In breast cancer screening, is it more important to minimize false negatives or false positives? Explain in 2–3 sentences, and state which metric (sensitivity or specificity) you would prioritize.

> *Your answer here*

### Q4.3

Find the optimal threshold for the Naïve Bayes model using Youden's J. Apply it and show the new confusion matrix. How did sensitivity change compared to the default 0.5 threshold?

```{r q4-3}
# Your code here

```

---

## Part 5 — PCA

### Q5.1

Run PCA on the training features (remember to scale). How many components are needed to explain at least 80% of the variance? Show the scree plot.

```{r q5-1}
# Your code here

```

### Q5.2

Show the loadings for the first two components. Which original variables contribute most to PC1? Can you give PC1 an interpretive name?

```{r q5-2}
# Your code here

```

### Q5.3

Plot the first two principal components colored by outcome. Is there visible separation between benign and malignant? How does this compare to the Pima diabetes PCA plot from class?

```{r q5-3}
# Your code here

```

---

## Part 6 — PCA + Classifier

### Q6.1

Using the number of components you chose in Q5.1, build a decision tree and a Naïve Bayes model on the PCA-transformed data. Report the AUC for each.

```{r q6-1}
# Your code here

```

### Q6.2

Compare all four models (raw tree, raw NB, PCA tree, PCA NB) in a single summary table with columns: Model, AUC, Number of Features. Did PCA help or hurt performance on this dataset?

```{r q6-2}
# Your code here

```

### Q6.3

In 3–4 sentences: when would you recommend using PCA as a preprocessing step before classification, and when would you skip it?

> *Your answer here*

---

*Good luck!*
