---
title: "Classification & Dimensionality Reduction"
subtitle: "Decision Trees · Naïve Bayes · PCA"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 9,
  fig.height = 5,
  fig.align  = "center"
)
```

---

## Setup

```{r packages}
# install.packages(c("mlbench", "rpart", "rpart.plot", "e1071",
#                     "caret", "pROC", "ggplot2", "dplyr",
#                     "corrplot", "factoextra"))

library(mlbench)
library(rpart)
library(rpart.plot)
library(e1071)
library(caret)
library(pROC)
library(ggplot2)
library(dplyr)
library(corrplot)
library(factoextra)

theme_set(theme_minimal(base_size = 14))
```

---

## The Dataset

We use the **Pima Indians Diabetes** dataset from the `mlbench` package: 768 women, 8 clinical features, binary outcome (diabetes yes/no).

```{r load-data}
data(PimaIndiansDiabetes)
df <- PimaIndiansDiabetes

dim(df)
str(df)
```

### Quick look

```{r summary}
summary(df)
```

### Outcome distribution

35% prevalence — this is an **imbalanced** dataset. Keep this in mind when we evaluate models later.

```{r outcome-dist}
table(df$diabetes)
prop.table(table(df$diabetes))

ggplot(df, aes(x = diabetes, fill = diabetes)) +
  geom_bar(width = 0.5) +
  scale_fill_manual(values = c("neg" = "#5eead4", "pos" = "#fb7185")) +
  labs(title = "Outcome Distribution", x = NULL, y = "Count") +
  theme(legend.position = "none")
```

### Explore features

```{r feature-distributions, fig.height=8}
df %>%
  tidyr::pivot_longer(-diabetes, names_to = "feature", values_to = "value") %>%
  ggplot(aes(x = value, fill = diabetes)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("neg" = "#5eead4", "pos" = "#fb7185")) +
  facet_wrap(~feature, scales = "free", ncol = 2) +
  labs(title = "Feature Distributions by Outcome") +
  theme(legend.position = "top")
```

> Which features seem to separate the two groups visually? Glucose, mass, and age stand out.

### Train/test split

We hold out 25% for testing. All model comparisons use the **same** split.

```{r split}
set.seed(42)
idx   <- createDataPartition(df$diabetes, p = 0.75, list = FALSE)
train <- df[idx, ]
test  <- df[-idx, ]

cat("Train:", nrow(train), "| Test:", nrow(test), "\n")
```

---

## Part 1 — Decision Trees

### What is classification?

**Classification** = given a set of measurements, predict which *category* an observation belongs to.

Clinical examples: diabetes (pos/neg), risk level (low/high), tumor (benign/malignant).

### What is a decision tree?

A series of if/else questions on the data, learned automatically from examples. Each path through the tree ends in a prediction.

### Build a first tree

```{r tree-fit}
tree_model <- rpart(diabetes ~ ., data = train, method = "class")

tree_model
```

### Visualize it

```{r tree-plot, fig.width=10, fig.height=7}
rpart.plot(tree_model,
           type     = 4,
           extra    = 104,
           fallen.leaves = TRUE,
           roundint = FALSE,
           cex      = 0.9,
           main     = "Decision Tree — Pima Diabetes")
```

> **Reading the tree:** Start at the top node. At each split, go left if the condition is TRUE. The leaf shows the predicted class, the probability, and the percentage of training data that ends up there.

### How does it choose where to split? — Gini Impurity

At each node, the algorithm tries every possible split and picks the one that best separates the classes. It measures "how mixed" a node is using **Gini Impurity**:

$$Gini = 1 - \sum_{i} p_i^2$$

Where $p_i$ = proportion of class $i$ in the node.

- Gini = 0.0 → **pure** node (all one class)
- Gini = 0.5 → **maximum mix** (50/50 split)

```{r gini-demo}
# Manual Gini calculation for a node with 70% neg, 30% pos
p_neg <- 0.70
p_pos <- 0.30
gini  <- 1 - (p_neg^2 + p_pos^2)
cat("Gini for 70/30 split:", round(gini, 3), "\n")

# Pure node
cat("Gini for 100/0 split:", 1 - (1^2 + 0^2), "\n")

# Maximum mix
cat("Gini for 50/50 split:", 1 - (0.5^2 + 0.5^2), "\n")
```

### Variable importance

Which features did the tree rely on most?

```{r var-importance}
vi <- tree_model$variable.importance
barplot(sort(vi), horiz = TRUE, las = 1, col = "#5eead4",
        main = "Variable Importance (Decision Tree)")
```

### Overfitting, pruning, and `cp`

**Overfitting** — the tree learns the training data so well that it fails on new patients.

**Pruning** — cutting back branches to keep the tree simpler.

**`cp`** (complexity parameter) — controls how aggressively we prune. Higher cp → simpler tree.

```{r cp-table}
printcp(tree_model)
```

```{r cp-plot}
plotcp(tree_model)
```

> **Rule of thumb:** Pick the cp value that gives the lowest cross-validated error (leftmost point below the dashed line).

### What does an overfit tree look like?

Let's build one with a tiny cp and see what happens:

```{r deep-tree, fig.width=12, fig.height=8}
tree_deep <- rpart(diabetes ~ ., data = train, method = "class",
                   control = rpart.control(cp = 0.001))
rpart.plot(tree_deep, cex = 0.55)
```

> The tree is huge and unreadable — classic overfitting. It memorized the training data.

### Build the pruned tree

```{r pruned-tree, fig.width=10, fig.height=7}
best_cp <- tree_model$cptable[which.min(tree_model$cptable[, "xerror"]), "CP"]
cat("Best cp:", best_cp, "\n")

tree_pruned <- prune(tree_model, cp = best_cp)

rpart.plot(tree_pruned,
           type     = 4,
           extra    = 104,
           fallen.leaves = TRUE,
           roundint = FALSE,
           cex      = 0.9,
           main     = "Pruned Decision Tree")
```

### Tree predictions

```{r tree-predict}
tree_pred       <- predict(tree_pruned, test, type = "class")
tree_pred_probs <- predict(tree_pruned, test, type = "prob")[, "pos"]
```

### Advantages & disadvantages

| ✅ Advantages | ❌ Disadvantages |
|:---|:---|
| Intuitive — easy to explain | Prone to overfitting |
| No feature scaling needed | Unstable — small data changes → different tree |
| Handles mixed data types | Axis-aligned splits only |

---

## Part 2 — Naïve Bayes

### From labels to probabilities

Unlike a decision tree that gives yes/no, Naïve Bayes gives a **probability**.
It uses Bayes' theorem to update a prior belief with new evidence.

### Bayes' Theorem

$$P(\text{Disease} \mid \text{Symptoms}) = \frac{P(\text{Symptoms} \mid \text{Disease}) \times P(\text{Disease})}{P(\text{Symptoms})}$$

- **Posterior** $P(\text{Disease} \mid \text{Symptoms})$ — what we want to know
- **Likelihood** $P(\text{Symptoms} \mid \text{Disease})$ — how common are these symptoms in sick patients
- **Prior** $P(\text{Disease})$ — baseline disease rate (35% in our data)

### Why "Naïve"?

It assumes every feature is **statistically independent** of every other — e.g., that glucose and BMI don't affect each other.

This is almost never true in clinical data, but the method still performs surprisingly well.

Let's check how correlated our features actually are:

```{r correlation, fig.width=8, fig.height=7}
cor_mat <- cor(train[, -9])
corrplot(cor_mat, method = "color", type = "upper", addCoef.col = "black",
         tl.col = "black", number.cex = 0.8,
         title = "Feature Correlations", mar = c(0, 0, 2, 0))
```

> Some features are clearly correlated (e.g., age & pregnant, insulin & glucose). The "naïve" assumption ignores this — yet the model often works well anyway.

### Build a Naïve Bayes model

```{r nb-fit}
nb_model <- naiveBayes(diabetes ~ ., data = train)

nb_model
```

> Each feature shows the mean and standard deviation **per class**. The model assumes each feature follows a normal (Gaussian) distribution within each class.

### Predict on a new patient

Glucose = 180, BMI = 33, Age = 45 — high-risk profile.

```{r nb-new-patient}
new_patient <- data.frame(
  pregnant = 3,
  glucose  = 180,
  pressure = 72,
  triceps  = 30,
  insulin  = 200,
  mass     = 33,
  pedigree = 0.5,
  age      = 45
)

predict(nb_model, new_patient, type = "raw")   # probabilities
predict(nb_model, new_patient, type = "class")  # class label
```

### And a low-risk patient

```{r nb-healthy-patient}
healthy_patient <- data.frame(
  pregnant = 0,
  glucose  = 80,
  pressure = 60,
  triceps  = 20,
  insulin  = 50,
  mass     = 21,
  pedigree = 0.2,
  age      = 22
)

predict(nb_model, healthy_patient, type = "raw")
predict(nb_model, healthy_patient, type = "class")
```

> Very low probability of diabetes — makes clinical sense.

### Naïve Bayes predictions on test set

```{r nb-predict}
nb_pred       <- predict(nb_model, test, type = "class")
nb_pred_probs <- predict(nb_model, test, type = "raw")[, "pos"]
```

### Advantages & disadvantages

| ✅ Advantages | ❌ Disadvantages |
|:---|:---|
| Extremely fast | Independence assumption rarely holds |
| Works with small training sets | Probability calibration can be off |
| Outputs probabilities, not just labels | Struggles with correlated features |

---

## Part 3 — Model Evaluation

### The accuracy trap

```{r accuracy-trap}
# What if we always predict "neg"?
always_neg <- factor(rep("neg", nrow(test)), levels = c("neg", "pos"))
mean(always_neg == test$diabetes)
```

> A model that **always predicts "negative"** achieves ~65% accuracy — because 65% of patients are actually negative. It learned nothing useful.

### Confusion matrix — full definitions

|  | **Predicted Neg** | **Predicted Pos** |
|:---|:---:|:---:|
| **Actual Neg** | TN (True Negative) | FP (False Positive) |
| **Actual Pos** | FN (False Negative) | TP (True Positive) |

- **True Positive (TP):** Sick patient correctly identified as sick
- **True Negative (TN):** Healthy patient correctly identified as healthy
- **False Positive (FP):** Healthy patient incorrectly flagged as sick (false alarm)
- **False Negative (FN):** Sick patient incorrectly cleared as healthy (missed case)

### Clinical priority: FN > FP

**Missing a sick patient (FN) is worse than a false alarm (FP).**

In clinical settings, we usually prioritize catching true positives — even at the cost of some false alarms.

### Decision Tree — confusion matrix & metrics

```{r tree-cm}
confusionMatrix(tree_pred, test$diabetes, positive = "pos")
```

### Naïve Bayes — confusion matrix & metrics

```{r nb-cm}
confusionMatrix(nb_pred, test$diabetes, positive = "pos")
```

### Key metrics — defined

- **Sensitivity (Recall)** = TP / (TP + FN) — of all *actually sick* patients, how many did we catch?

- **Specificity** = TN / (TN + FP) — of all *actually healthy* patients, how many did we correctly clear?

- **AUC** (Area Under the ROC Curve) = overall ability to distinguish sick from healthy. Ranges from 0 to 1; 0.5 = coin flip.

### The ROC Curve

**ROC** = Receiver Operating Characteristic. It plots sensitivity (y-axis) vs. false positive rate (x-axis) at every possible classification threshold.

**AUC** = the area under this curve. Higher = better discrimination between classes.

```{r roc-curves, fig.width=8, fig.height=6}
roc_tree <- roc(test$diabetes, tree_pred_probs, levels = c("neg", "pos"))
roc_nb   <- roc(test$diabetes, nb_pred_probs,   levels = c("neg", "pos"))

plot(roc_tree, col = "#5eead4", lwd = 2,
     main = "ROC Curves — Decision Tree vs. Naïve Bayes",
     legacy.axes = TRUE, print.auc = TRUE, print.auc.y = 0.4)
plot(roc_nb, col = "#fb7185", lwd = 2, add = TRUE,
     print.auc = TRUE, print.auc.y = 0.3)
legend("bottomright",
       legend = c(paste0("Decision Tree (AUC=", round(auc(roc_tree), 3), ")"),
                  paste0("Naïve Bayes (AUC=",   round(auc(roc_nb), 3),   ")")),
       col = c("#5eead4", "#fb7185"), lwd = 2)
```

### Threshold tuning

The default threshold is 0.5, but we can change it to favor sensitivity over specificity.

**Youden's J** finds the threshold that maximizes sensitivity + specificity simultaneously.

```{r threshold-tuning}
coords_tree <- coords(roc_tree, "best", best.method = "youden")
coords_nb   <- coords(roc_nb,   "best", best.method = "youden")

cat("Decision Tree — optimal threshold:", round(coords_tree$threshold, 3),
    "→ Sensitivity:", round(coords_tree$sensitivity, 3),
    ", Specificity:", round(coords_tree$specificity, 3), "\n")

cat("Naïve Bayes  — optimal threshold:", round(coords_nb$threshold, 3),
    "→ Sensitivity:", round(coords_nb$sensitivity, 3),
    ", Specificity:", round(coords_nb$specificity, 3), "\n")
```

### Apply the optimal threshold to Naïve Bayes

```{r nb-tuned}
optimal_threshold <- coords_nb$threshold
nb_pred_tuned     <- factor(ifelse(nb_pred_probs >= optimal_threshold, "pos", "neg"),
                            levels = c("neg", "pos"))
confusionMatrix(nb_pred_tuned, test$diabetes, positive = "pos")
```

> Sensitivity increased (fewer FN), specificity decreased (more FP) — exactly the tradeoff we expected.

---

## ☕ Break

---

## Part 4 — PCA

### Principal Component Analysis — what is it?

**PCA** (Principal Component Analysis) takes many correlated variables and reduces them to a smaller set of new, uncorrelated variables called **principal components** — while keeping as much of the original information as possible.

### The Curse of Dimensionality — why we need PCA

More variables → more problems:

- Can't visualize beyond 3D
- Need exponentially more data to fill the space
- Correlated features = redundancy
- Irrelevant features add noise

### Prepare the data

PCA works on **numeric features only**, and we must **standardize** (scale) them first.

```{r pca-prep}
features <- train[, -9]
outcome  <- train[, 9]

features_test <- test[, -9]
outcome_test  <- test[, 9]
```

### Why `scale = TRUE`?

PCA is driven by **variance** (how spread out the data is). Without standardizing, variables with larger numeric ranges dominate.

```{r scale-demo}
apply(features, 2, range)
```

> Glucose ranges 0–199, while pedigree ranges 0.08–2.42. Without scaling, glucose would dominate PC1 simply because its numbers are bigger, not because it's more informative.

### Run PCA

```{r pca-fit}
pca <- prcomp(features, center = TRUE, scale. = TRUE)

summary(pca)
```

### Variance explained — the scree plot

A **scree plot** shows the variance explained by each component, in descending order. Look for where the curve flattens — adding more components past that point stops helping.

```{r scree, fig.width=8, fig.height=5}
fviz_eig(pca, addlabels = TRUE, ylim = c(0, 40),
         main = "Scree Plot — Variance Explained by Each Component",
         barfill = "#22d3ee", barcolor = "#1e293b")
```

**Rule of thumb:** Keep enough components to explain ≥ 80% of total variance.

```{r cumvar}
cumvar <- cumsum(pca$sdev^2 / sum(pca$sdev^2)) * 100
data.frame(PC = paste0("PC", 1:8), Cumulative_Variance_Pct = round(cumvar, 1))
```

### Loadings — what do the components mean?

Each principal component is a weighted combination of the original variables. These weights are called **loadings**.

- High loading = that variable contributes strongly to that component
- Example: if glucose, insulin, and age all load heavily on PC1, you might interpret it as a "metabolic risk" axis

```{r loadings-plot, fig.width=9, fig.height=6}
fviz_pca_var(pca,
             col.var    = "contrib",
             gradient.cols = c("#475569", "#22d3ee", "#5eead4"),
             repel      = TRUE,
             title      = "PCA — Variable Loadings (Contributions)")
```

```{r loading-table}
round(pca$rotation[, 1:3], 3)
```

### PCA biplot — observations colored by outcome

```{r biplot, fig.width=9, fig.height=7}
pca_scores <- as.data.frame(pca$x)
pca_scores$diabetes <- outcome

ggplot(pca_scores, aes(x = PC1, y = PC2, color = diabetes)) +
  geom_point(alpha = 0.6, size = 2) +
  scale_color_manual(values = c("neg" = "#5eead4", "pos" = "#fb7185")) +
  labs(title = "PCA — First Two Components",
       x = paste0("PC1 (", round(cumvar[1], 1), "%)"),
       y = paste0("PC2 (", round(cumvar[2] - cumvar[1], 1), "%)")) +
  theme(legend.position = "top")
```

> Can you see separation between the two groups? It's partial — these two components don't capture everything.

### PC1 vs PC3

Does a different component pair show better separation?

```{r pc1-pc3, fig.width=9, fig.height=7}
ggplot(pca_scores, aes(x = PC1, y = PC3, color = diabetes)) +
  geom_point(alpha = 0.6, size = 2) +
  scale_color_manual(values = c("neg" = "#5eead4", "pos" = "#fb7185")) +
  labs(title = "PC1 vs PC3",
       x = paste0("PC1 (", round(cumvar[1], 1), "%)"),
       y = paste0("PC3 (", round(cumvar[3] - cumvar[2], 1), "%)")) +
  theme(legend.position = "top")
```

> PC3 explains less variance than PC2, so separation is usually not better.

---

## Part 5 — PCA → Classifier

### The idea

**Reduce dimensions first, then classify.** Feed PCA components into a decision tree or Naïve Bayes instead of the raw features. Does performance change?

### Prepare PCA-transformed data

```{r pca-transform}
k <- which(cumvar >= 80)[1]
cat("Keeping", k, "components (", round(cumvar[k], 1), "% variance)\n")

train_pca <- as.data.frame(predict(pca, features)[, 1:k])
train_pca$diabetes <- outcome

test_pca <- as.data.frame(predict(pca, features_test)[, 1:k])
test_pca$diabetes <- outcome_test
```

### Decision Tree on PCA components

```{r tree-pca}
tree_pca <- rpart(diabetes ~ ., data = train_pca, method = "class")

tree_pca_pred  <- predict(tree_pca, test_pca, type = "class")
tree_pca_probs <- predict(tree_pca, test_pca, type = "prob")[, "pos"]

confusionMatrix(tree_pca_pred, test_pca$diabetes, positive = "pos")
```

### Naïve Bayes on PCA components

```{r nb-pca}
nb_pca <- naiveBayes(diabetes ~ ., data = train_pca)

nb_pca_pred  <- predict(nb_pca, test_pca, type = "class")
nb_pca_probs <- predict(nb_pca, test_pca, type = "raw")[, "pos"]

confusionMatrix(nb_pca_pred, test_pca$diabetes, positive = "pos")
```

### What about only 2 components?

How much performance do we lose if we compress aggressively?

```{r pca-2-components}
train_pca2 <- as.data.frame(predict(pca, features)[, 1:2])
train_pca2$diabetes <- outcome
test_pca2  <- as.data.frame(predict(pca, features_test)[, 1:2])
test_pca2$diabetes  <- outcome_test

tree_2pc <- rpart(diabetes ~ ., data = train_pca2, method = "class")
nb_2pc   <- naiveBayes(diabetes ~ ., data = train_pca2)

roc_t2 <- roc(test_pca2$diabetes,
              predict(tree_2pc, test_pca2, type = "prob")[, "pos"],
              levels = c("neg", "pos"))
roc_n2 <- roc(test_pca2$diabetes,
              predict(nb_2pc, test_pca2, type = "raw")[, "pos"],
              levels = c("neg", "pos"))

cat("Tree (2 PCs) AUC:", round(auc(roc_t2), 3), "\n")
cat("NB   (2 PCs) AUC:", round(auc(roc_n2), 3), "\n")
```

> Performance drops — 2 PCs aren't enough to capture 80% variance here.

### Compare all four models — ROC curves

```{r roc-all, fig.width=9, fig.height=7}
roc_tree_pca <- roc(test_pca$diabetes, tree_pca_probs, levels = c("neg", "pos"))
roc_nb_pca   <- roc(test_pca$diabetes, nb_pca_probs,   levels = c("neg", "pos"))

plot(roc_tree, col = "#5eead4", lwd = 2.5,
     main = "ROC Comparison — Raw Features vs. PCA",
     legacy.axes = TRUE)
plot(roc_nb,       col = "#fb7185", lwd = 2.5, add = TRUE)
plot(roc_tree_pca, col = "#22d3ee", lwd = 2, lty = 2, add = TRUE)
plot(roc_nb_pca,   col = "#fbbf24", lwd = 2, lty = 2, add = TRUE)

legend("bottomright",
       legend = c(
         paste0("Tree — Raw (AUC=",      round(auc(roc_tree), 3),     ")"),
         paste0("NB — Raw (AUC=",        round(auc(roc_nb), 3),       ")"),
         paste0("Tree — PCA (AUC=",      round(auc(roc_tree_pca), 3), ")"),
         paste0("NB — PCA (AUC=",        round(auc(roc_nb_pca), 3),   ")")
       ),
       col = c("#5eead4", "#fb7185", "#22d3ee", "#fbbf24"),
       lwd = 2, lty = c(1, 1, 2, 2))
```

### Summary table

```{r summary-table}
results <- data.frame(
  Model = c("Decision Tree", "Naïve Bayes", "Tree + PCA", "NB + PCA"),
  AUC   = c(round(auc(roc_tree), 3),
            round(auc(roc_nb), 3),
            round(auc(roc_tree_pca), 3),
            round(auc(roc_nb_pca), 3)),
  Features = c(8, 8, k, k)
)
results
```

> **Similar AUC, fewer features.** PCA didn't hurt performance — and reduced complexity. With higher-dimensional data (hundreds or thousands of features), the gain from PCA is much larger.

### Side-by-side: same patient, all four models

```{r side-by-side}
new_patient <- data.frame(
  pregnant = 3, glucose = 180, pressure = 72, triceps = 30,
  insulin = 200, mass = 33, pedigree = 0.5, age = 45
)

cat("=== Raw Features ===\n")
cat("Decision Tree:", predict(tree_pruned, new_patient, type = "prob")[, "pos"], "\n")
cat("Naïve Bayes:  ", predict(nb_model, new_patient, type = "raw")[, "pos"], "\n")

new_patient_pca <- as.data.frame(predict(pca, new_patient))[, 1:k, drop = FALSE]
cat("\n=== PCA Features ===\n")
cat("Tree + PCA:   ", predict(tree_pca, new_patient_pca, type = "prob")[, "pos"], "\n")
cat("NB + PCA:     ", predict(nb_pca, new_patient_pca, type = "raw")[, "pos"], "\n")
```

---

## Summary

| Method | Best for | Less suitable |
|:---|:---|:---|
| **Decision Tree** | Interpretability, first exploration | Noisy data |
| **Naïve Bayes** | Small samples, probability estimates | Correlated features |
| **PCA** | Many variables, preprocessing | Few variables |
| **PCA + Classifier** | High-dimensional, multicollinearity | Clean data, few features |

### Key takeaways

1. **Trees = Transparency** — interpretable, explain to clinicians
2. **Bayes = Probabilities** — gives calibrated (ish) risk scores
3. **PCA = Compression** — fewer features, less noise, faster models
4. **Accuracy ≠ enough** — always look at sensitivity, specificity, and AUC

---

