---
title: "Homework Solution — Publication-Ready Plots"
subtitle: "Week 10 · Part II"
author: "R for Clinical Research"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.width = 7, fig.height = 4.5, dpi = 150
)
```

```{r libraries}
library(ggplot2)
library(dplyr)
library(tidyr)
library(forcats)
library(ggpubr)
library(patchwork)
library(mlbench)
library(rcartocolor)

data(PimaIndiansDiabetes2)
```

---

## Q1 — Prepare the data

```{r q1}
n_start <- nrow(PimaIndiansDiabetes2)

pima <- PimaIndiansDiabetes2 |>
  drop_na(mass, glucose, age, diabetes) |>
  mutate(
    age_group = cut(
      age,
      breaks = c(20, 30, 40, 50, Inf),
      labels = c("21-30", "31-40", "41-50", "51+"),
      right  = TRUE
    )
  )

n_dropped <- n_start - nrow(pima)
n_dropped
nrow(pima)

count(pima, age_group)
count(pima, age_group, diabetes)
```

`r n_dropped` rows dropped (mostly missing `mass`), `r nrow(pima)` remain.

**Bias note:** the missingness is not random — a woman with no recorded BMI or
glucose is likely to differ systematically from one who was fully measured
(different clinic visit, different reason for testing). Dropping her removes a
non-random slice of the cohort, so every prevalence in the figure is a
prevalence *among the completely measured*, not among the cohort.

---

## Q2 — Your theme

```{r q2}
theme_hw <- function(base_size = 12) {
  theme_classic(base_size = base_size) +
    theme(
      text            = element_text(family = "Helvetica", color = "grey20"),
      axis.line       = element_line(color = "black", linewidth = 0.4),
      axis.ticks      = element_line(color = "black", linewidth = 0.4),
      panel.grid      = element_blank(),
      plot.title      = element_text(face = "bold", size = base_size + 1),
      plot.subtitle   = element_text(color = "grey40"),
      legend.position = "bottom",
      legend.title    = element_blank()
    )
}
```

---

## Q3 — Panel A: distribution

```{r q3, fig.height=3.5}
p_hist <- ggplot(pima, aes(mass)) +
  geom_histogram(
    aes(y = after_stat(density)),
    binwidth  = 2,
    fill      = "#94a3b8",
    color     = "white",
    linewidth = 0.2
  ) +
  geom_density(color = "#0f172a", linewidth = 0.6) +
  labs(x = "BMI (kg/m²)", y = "Density") +
  theme_hw()

p_hist
```

**Binwidth justification:** BMI in this cohort spans roughly 18–67 kg/m²; a
2 kg/m² bin gives ~25 bins — enough to show the right skew without turning the
histogram into noise, and it is close to the clinical granularity people
actually act on (BMI categories are 5-unit wide).

---

## Q4 — Panel B: BMI by age group and diabetes status

```{r q4, fig.height=4}
p_box <- ggplot(pima, aes(age_group, mass, fill = diabetes)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA, width = 0.6) +
  geom_point(
    position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.6),
    alpha = 0.3, size = 0.9, color = "grey30"
  ) +
  stat_compare_means(
    aes(group = diabetes),
    method = "wilcox.test",
    label  = "p.signif",
    size   = 3
  ) +
  scale_fill_carto_d(palette = "Safe") +
  labs(x = "Age group (years)", y = "BMI (kg/m²)") +
  theme_hw()

p_box
```

---

## Q5 — Panel C: prevalence

```{r q5, fig.height=3.5}
prev <- pima |>
  group_by(age_group) |>
  summarise(
    n    = n(),
    pos  = sum(diabetes == "pos"),
    pct  = 100 * pos / n,
    .groups = "drop"
  )

prev

p_bar <- ggplot(prev, aes(age_group, pct, fill = age_group)) +
  geom_col(width = 0.65) +
  geom_text(aes(label = sprintf("%.1f%%", pct)),
            vjust = -0.5, size = 3.2, family = "Helvetica") +
  scale_fill_carto_d(palette = "Safe") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(x = "Age group (years)", y = "Diabetes prevalence (%)") +
  theme_hw() +
  theme(legend.position = "none")

p_bar
```

**Why not `fct_reorder`:** age group is an **ordinal** variable. Sorting by
value would destroy the axis the reader needs in order to see the trend —
the whole point of the panel is that prevalence rises with age and then
drops back in the oldest group. Reorder only nominal categories.

---

## Q6 — Assemble and export

```{r q6, fig.height=6.5, fig.width=8}
final <- p_hist / (p_box + p_bar) +
  plot_annotation(
    title    = "Figure 1. BMI and diabetes prevalence across age groups in the Pima cohort",
    subtitle = "(A) Distribution of BMI. (B) BMI by age group and diabetes status. (C) Diabetes prevalence by age group.",
    tag_levels = "A",
    theme = theme(
      plot.title    = element_text(face = "bold", size = 13, family = "Helvetica"),
      plot.subtitle = element_text(color = "grey40", size = 10, family = "Helvetica")
    )
  ) +
  plot_layout(guides = "collect", heights = c(1, 1.2)) &
  theme(legend.position = "bottom")

final
```

```{r q6-save, eval=FALSE}
ggsave(
  filename = "figure_hw_pima.pdf",
  plot     = final,
  width    = 7,
  height   = 6,
  units    = "in",
  dpi      = 300,
  device   = cairo_pdf
)
```

---

## Q7 — Peer review

Typical failures on first render, and the fix:

| Failed item | Fix |
|---|---|
| Panel B legend duplicated / floating | `plot_layout(guides = "collect")` + `& theme(legend.position = "bottom")` |
| Significance stars clipped at the top of B | `scale_y_continuous(expand = expansion(mult = c(0.05, 0.12)))` |
| "51+" bar label cut off | Y expansion `mult = c(0, 0.15)` on Panel C |
| Panels A and C use different greys for the same thing | They encode different variables — keep, but never reuse a colour across panels for different meanings |
| Text unreadable at 100% | Drop to `theme_hw(base_size = 11)` rather than shrinking the figure |

---

# Grading rubric

| Q | Points | Full credit requires |
|---|---|---|
| Q1 | 10 | correct `cut()` breaks/labels, counts reported, bias sentence names *non-random* missingness |
| Q2 | 10 | a **function** with `base_size`, all six theme elements, reused everywhere |
| Q3 | 15 | density scale (`after_stat(density)`) so the curve overlays correctly, binwidth justified, units on axes |
| Q4 | 20 | `position_jitterdodge`, `aes(group = diabetes)` in the stat, colorblind-safe palette |
| Q5 | 20 | percentage not count, age order preserved **and** justified, labels not clipped |
| Q6 | 20 | `p_hist / (p_box + p_bar)`, tags, 300 DPI, `cairo_pdf` |
| Q7 | 5 | at least two concrete failures with the fix applied |

Common deductions: default `theme_grey()` sneaking into one panel; axis labels
without units; `geom_jitter()` instead of `position_jitterdodge()` in Q4
(points detach from their box); counts instead of percentages in Q5.
