---
title: "Homework — Publication-Ready Plots"
subtitle: "Week 10 · Part II"
author: "Your name"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.align = "center", fig.width = 7, fig.height = 4.5, dpi = 150
)
```

# Setup

```{r libraries}
library(ggplot2)
library(dplyr)
library(tidyr)
library(forcats)
library(ggpubr)
library(patchwork)
library(mlbench)
library(rcartocolor)

data(PimaIndiansDiabetes2)
```

The deliverable is **one PDF figure**, 300 DPI, three panels, one story.
The story this time is **BMI**, not glucose.

---

## Q1 — Prepare the data (10 pts)

Starting from `PimaIndiansDiabetes2`:

a. Keep only rows with complete `mass`, `glucose`, `age`, `diabetes`.
b. Create `age_group`: a factor with levels `"21-30"`, `"31-40"`, `"41-50"`, `"51+"`.
c. Report how many rows you dropped and how many remain per `age_group`.

**Write one sentence** on why dropping incomplete rows here could bias the figure.

```{r q1}
# your code

```

---

## Q2 — Your theme (10 pts)

Write `theme_hw()` — a function with a `base_size` argument, built on
`theme_classic()`, with: no panel grid, bold title, black axis lines,
legend at the bottom, no legend title.

Every plot below must use it.

```{r q2}
# your code

```

---

## Q3 — Panel A: distribution (15 pts)

Histogram of `mass` with a density overlay.

- Choose a `binwidth` **and justify it in one sentence**.
- Axis labels with units.
- No legend.

```{r q3}
# your code

```

---

## Q4 — Panel B: BMI by age group and diabetes status (20 pts)

Boxplot of `mass` by `age_group`, filled by `diabetes`, with jittered points
(use `position_jitterdodge()` so points land on their own box).

- Colorblind-safe fill palette.
- Add a Wilcoxon test comparing `neg` vs `pos` **within each age group**
  (hint: `stat_compare_means(aes(group = diabetes), label = "p.signif")`).

```{r q4}
# your code

```

---

## Q5 — Panel C: prevalence, not counts (20 pts)

For each `age_group`, compute the **percentage** of patients with `diabetes == "pos"`.
Plot it as a bar chart.

- Bars in age order (not sorted by value — order matters here; say why in one sentence).
- Percentage printed above each bar, one decimal.
- Y axis expanded so labels are not clipped.

```{r q5}
# your code

```

---

## Q6 — Assemble and export (20 pts)

- Layout: A on top, B and C on the bottom row.
- `plot_annotation()` with a title, a subtitle describing each panel, and tags A/B/C.
- Save as `figure_hw_pima.pdf`, 7 × 6 in, 300 DPI, `cairo_pdf`.

```{r q6}
# your code

```

---

## Q7 — Peer review (5 pts)

Open your PDF at 100% zoom. Run the Part 04 checklist from the companion
notebook. List any item that failed and what you changed.

*your answer here*
