---
title: "Publication-Ready Plots in R"
subtitle: "Week 10 · Part II — Companion Notebook"
author: "R for Clinical Research"
date: "`r format(Sys.Date(), '%B %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 4.5,
  dpi       = 150
)
```

# Before you begin

This notebook is the hands-on companion to the **Publication-Ready Plots** lecture.
By the end you will produce **one PDF figure** at 300 DPI containing three coordinated
panels — the kind of figure a journal reviewer expects.

## Install the packages (one time)

```{r install, eval=FALSE}
install.packages(c(
  "ggplot2",     # the engine
  "dplyr",       # data wrangling
  "tidyr",       # drop_na
  "forcats",     # reordering factors
  "ggpubr",      # statistical annotations + publication themes
  "patchwork",   # combining plots
  "mlbench",     # the dataset for the project (Pima diabetes)
  "rcartocolor"  # colorblind-safe palettes
))
```

## Load everything

```{r libraries}
library(ggplot2)
library(dplyr)
library(tidyr)
library(forcats)
library(ggpubr)
library(patchwork)
library(mlbench)
library(rcartocolor)
```

## Meet the data

We use `PimaIndiansDiabetes2` — the corrected version, where physiologically
impossible zeros (a glucose of 0, a BMI of 0) are coded as `NA` instead of
being treated as real measurements.

```{r data-peek}
data(PimaIndiansDiabetes2)

pima <- PimaIndiansDiabetes2 |>
  drop_na(glucose, mass, age)   # keep rows complete on the variables we plot

glimpse(pima)
```

Each row is a woman of Pima heritage. The outcome `diabetes` has two levels
(`neg` / `pos`), and we have continuous clinical measurements: plasma glucose
(mg/dL), BMI (kg/m²), diastolic blood pressure (mm Hg), age (years), and more.

---

# Part 01 — The publication standard

## Default vs. polished — see the gap

Let's see how a default `ggplot` looks against a publication-ready version of
**the exact same data**.

### Default

```{r default-plot}
ggplot(pima, aes(diabetes, glucose, fill = diabetes)) +
  geom_boxplot()
```

This works. But: gray panel, gridlines competing with the data, default
ggplot palette, no units on the y-axis, redundant legend.

### Publication-ready

```{r polished-plot}
ggplot(pima, aes(diabetes, glucose)) +
  geom_boxplot(fill = "#2c3e50", alpha = 0.85, width = 0.5,
               outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.35, size = 1) +
  labs(x = NULL, y = "Plasma glucose (mg/dL)") +
  theme_classic(base_size = 12) +
  theme(text = element_text(family = "Helvetica"))
```

Same data, ten lines of code, very different first impression.

---

## Vector vs. raster — pick the right format

| Format | Type   | Use it for                              |
|--------|--------|-----------------------------------------|
| PDF    | Vector | Default for journal submission          |
| SVG    | Vector | Web, post-processing in Illustrator     |
| PNG    | Raster | Plots that include rendered images      |
| TIFF   | Raster | Required by some journals (Cell, Nature)|
| JPEG   | Raster | **Never** — lossy, smears thin lines    |

**Rule:** if the plot is mostly lines, text, and shapes → vector.
If it contains a rendered image (heatmap, microscopy) → raster at 300 DPI.

---

## Labels with units — the small detail that signals seriousness

```{r labels-good, fig.height=3.5}
ggplot(pima, aes(glucose, mass)) +
  geom_point(alpha = 0.5, color = "#22d3ee") +
  labs(
    title    = "BMI rises with plasma glucose",
    subtitle = "Pima Indian diabetes cohort",
    x        = "Plasma glucose (mg/dL)",
    y        = "BMI (kg/m²)",
    caption  = "Source: mlbench::PimaIndiansDiabetes2"
  ) +
  theme_classic(base_size = 12)
```

Every axis has a unit. Title is in sentence case. Subtitle and caption do
the job a paragraph would do.

---

## Colorblind-safe palettes

```{r palette-comparison, fig.height=3.5}
# Default — two random hues
p_default <- ggplot(pima, aes(diabetes, glucose, fill = diabetes)) +
  geom_boxplot() +
  labs(title = "Default palette") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")

# viridis — survives B&W and colorblind viewers
p_viridis <- ggplot(pima, aes(diabetes, glucose, fill = diabetes)) +
  geom_boxplot() +
  scale_fill_viridis_d(option = "D") +
  labs(title = "viridis") +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none")

p_default + p_viridis
```

For categorical variables, prefer `scale_fill_viridis_d()`,
`scale_fill_brewer(palette = "Set2")`, or `rcartocolor::scale_fill_carto_d(palette = "Safe")`.

---

# Part 02 — Advanced R toolbox

## A reusable theme

Build it once. Save it. Apply it to every plot in the paper.

```{r my-theme}
my_theme <- theme_classic(base_size = 12) +
  theme(
    text             = element_text(family = "Helvetica", color = "grey20"),
    axis.line        = element_line(color = "black", linewidth = 0.4),
    axis.ticks       = element_line(color = "black", linewidth = 0.4),
    panel.grid       = element_blank(),
    plot.title       = element_text(face = "bold", size = 13),
    plot.subtitle    = element_text(color = "grey40"),
    strip.background = element_rect(fill = "grey90", color = NA),
    strip.text       = element_text(face = "bold"),
    legend.position  = "bottom",
    legend.title     = element_blank()
  )
```

Test it:

```{r my-theme-test, fig.height=3.5}
ggplot(pima, aes(glucose, mass, color = diabetes)) +
  geom_point(alpha = 0.7) +
  scale_color_carto_d(palette = "Safe") +
  labs(x = "Plasma glucose (mg/dL)", y = "BMI (kg/m²)") +
  my_theme
```

---

## Statistical annotations with `ggpubr`

`stat_compare_means()` adds p-values or significance stars without you
running the test by hand and pasting the number.

```{r stat-compare, fig.height=4}
# Significance stars (***, **, *, ns)
ggplot(pima, aes(diabetes, glucose, fill = diabetes)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.15, alpha = 0.3, size = 1) +
  stat_compare_means(
    comparisons = list(c("neg", "pos")),
    method = "wilcox.test",
    label  = "p.signif"
  ) +
  scale_fill_carto_d(palette = "Safe") +
  labs(x = NULL, y = "Plasma glucose (mg/dL)") +
  my_theme +
  theme(legend.position = "none")
```

To show numeric p-values instead of stars: `label = "p.format"`.

With only two groups, one comparison is enough. If you had three or more
groups, add every pairwise `c()` to `comparisons`, or drop `comparisons`
entirely and use an overall test (`method = "kruskal.test"`).

---

## Combining plots with `patchwork`

Three operators do everything you need:

```{r patchwork-demo, fig.height=5}
p1 <- ggplot(pima, aes(glucose)) +
  geom_histogram(bins = 25, fill = "#22d3ee") +
  labs(x = "Plasma glucose (mg/dL)", y = "Count") + my_theme

p2 <- ggplot(pima, aes(diabetes, glucose, fill = diabetes)) +
  geom_boxplot(alpha = 0.6) +
  scale_fill_carto_d(palette = "Safe") +
  labs(x = NULL, y = "Plasma glucose (mg/dL)") +
  my_theme + theme(legend.position = "none")

p3 <- pima |>
  count(diabetes) |>
  ggplot(aes(fct_reorder(diabetes, n), n, fill = diabetes)) +
  geom_col() +
  scale_fill_carto_d(palette = "Safe") +
  labs(x = NULL, y = "Patients counted") +
  my_theme + theme(legend.position = "none")

# Side by side: +
# Stacked: /
# Top + bottom-row: p1 / (p2 + p3)
p1 / (p2 + p3) +
  plot_annotation(tag_levels = "A")
```

The `tag_levels = "A"` adds the A/B/C panel labels reviewers will reference.

---

# Part 03 — The mission

> Build a **3-panel PDF figure** from the Pima diabetes cohort.
> 300 DPI. Coordinated theme. Coordinated palette. One story.

We'll use plasma glucose as the variable of interest and diabetes status as
the grouping.

## Step 1 — The histogram

```{r mission-histogram, fig.height=3.5}
p_hist <- ggplot(pima, aes(glucose)) +
  geom_histogram(
    aes(y = after_stat(density)),
    binwidth = 5,                # justify: 5 mg/dL bins match reporting precision
    fill     = "#94a3b8",
    color    = "white",
    linewidth = 0.2
  ) +
  geom_density(color = "#0f172a", linewidth = 0.6) +
  labs(
    x = "Plasma glucose (mg/dL)",
    y = "Density"
  ) +
  my_theme

p_hist
```

**Publication checklist for this panel:**

- ☑ binwidth chosen on purpose, not by default
- ☑ density curve overlay shows shape clearly
- ☑ axis labels include units
- ☑ no legend (one variable, one color)

## Step 2 — The boxplot

```{r mission-boxplot, fig.height=3.5}
p_box <- ggplot(pima, aes(diabetes, glucose, fill = diabetes)) +
  geom_boxplot(alpha = 0.5, outlier.shape = NA, width = 0.5) +
  geom_jitter(width = 0.15, alpha = 0.35, size = 1, color = "grey30") +
  stat_compare_means(
    comparisons = list(c("neg", "pos")),
    method = "wilcox.test",
    label  = "p.signif",
    size   = 3
  ) +
  scale_fill_carto_d(palette = "Safe") +
  labs(x = NULL, y = "Plasma glucose (mg/dL)") +
  my_theme +
  theme(legend.position = "none")

p_box
```

**Why jitter instead of just a box:** the box hides the raw data. Jittered
points show the reader the sample size and how dense each group is.

## Step 3 — The bar chart

```{r mission-bars, fig.height=3.5}
p_bar <- pima |>
  count(diabetes) |>
  ggplot(aes(fct_reorder(diabetes, n), n, fill = diabetes)) +
  geom_col(width = 0.65) +
  geom_text(aes(label = n), vjust = -0.5, size = 3.5, family = "Helvetica") +
  scale_fill_carto_d(palette = "Safe") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(x = NULL, y = "Number of patients") +
  my_theme +
  theme(legend.position = "none")

p_bar
```

**The three publication moves:**

- `fct_reorder(diabetes, n)` — bars sorted by value, not alphabetically
- `geom_text(aes(label = n))` — values printed above each bar
- `legend.position = "none"` — the x-axis already names the groups

## Step 4 — Final assembly

```{r mission-final, fig.height=6.5, fig.width=8}
final <- p_hist / (p_box + p_bar) +
  plot_annotation(
    title    = "Figure 1. Plasma glucose and diabetes status in the Pima cohort",
    subtitle = "(A) Distribution of glucose. (B) Glucose by diabetes status. (C) Sample size per group.",
    tag_levels = "A",
    theme = theme(
      plot.title    = element_text(face = "bold", size = 13, family = "Helvetica"),
      plot.subtitle = element_text(color = "grey40", size = 10, family = "Helvetica")
    )
  ) +
  plot_layout(guides = "collect", heights = c(1, 1.2))

final
```

## Step 5 — Save it the right way

```{r save-pdf, eval=FALSE}
ggsave(
  filename = "figure_1_pima.pdf",
  plot     = final,
  width    = 7,        # inches — fits a double-column journal layout
  height   = 6,
  units    = "in",
  dpi      = 300,
  device   = cairo_pdf # ensures fonts embed cleanly
)
```

For a TIFF (some journals require it):

```{r save-tiff, eval=FALSE}
ggsave(
  filename    = "figure_1_pima.tiff",
  plot        = final,
  width       = 7,
  height      = 6,
  units       = "in",
  dpi         = 300,
  compression = "lzw"
)
```

---

# Part 04 — Peer review checklist

Open your saved PDF at 100% zoom and walk through this list:

- [ ] All text is readable at 100% — no need to zoom in
- [ ] Every axis has a label with units in parentheses
- [ ] Color is consistent across all panels (same group → same color)
- [ ] Figure works in black-and-white print (try grayscale view in your PDF reader)
- [ ] Panel tags A, B, C are present
- [ ] Saved at 300 DPI as PDF or TIFF — never JPEG
- [ ] Filename matches the manuscript reference (`figure_1_pima.pdf`)

If any item fails, fix the figure — not the manuscript text.

---

# Bonus — your own theme as a function

For long-term reuse, wrap the theme in a function:

```{r theme-function, eval=FALSE}
theme_paper <- function(base_size = 12) {
  theme_classic(base_size = base_size) +
    theme(
      text             = element_text(family = "Helvetica", color = "grey20"),
      axis.line        = element_line(color = "black", linewidth = 0.4),
      panel.grid       = element_blank(),
      plot.title       = element_text(face = "bold"),
      strip.background = element_rect(fill = "grey90", color = NA),
      legend.position  = "bottom",
      legend.title     = element_blank()
    )
}

# Use it:
ggplot(pima, aes(...)) + ... + theme_paper()

# Or with a different base size for slides:
ggplot(pima, aes(...)) + ... + theme_paper(base_size = 16)
```

Drop this into a small `R/themes.R` file in every project and `source()` it.
That is how labs achieve consistency across a whole paper.

---

# Resources

- **R Graph Gallery** — r-graph-gallery.com
- **From Data to Viz** — data-to-viz.com
- **ggplot2 book** (Wickham et al.) — ggplot2-book.org
- **Fundamentals of Data Visualization** (Claus Wilke) — clauswilke.com/dataviz
- **ggpubr** documentation — rpkgs.datanovia.com/ggpubr
- **patchwork** documentation — patchwork.data-imaginist.com
